### 2017-03-23 chap.tex
Initial commit

\documentclass{report}

\usepackage{amsmath}
\usepackage{txfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{url}

\title{Probabilistic Forecasting of the Disturbance Storm Time Index: An Autoregressive Gaussian Process approach}

\begin{document}

\begin{abstract}
We present a methodology for generating probabilistic predictions for the \emph{Disturbance Storm Time} ($Dst$) geomagnetic activity index. We focus on the \emph{One Step Ahead} (OSA) prediction task and use the OMNI hourly resolution data to build our models.

Our proposed methodology is based on the technique of \emph{Gaussian Process Regression} (GPR). Within this framework we develop two models; \emph{Gaussian Process Auto-Regressive} (GP-AR) and \emph{Gaussian Process Auto-Regressive with eXogenous inputs} (GP-ARX).

We also propose a criterion to aid model selection with respect to the order of auto-regressive inputs. Finally we test the performance of the GP-AR and GP-ARX models on a set of 63 geomagnetic storms between 1998 and 2006 and illustrate sample predictions with error bars for some of these events.

\end{abstract}

\section{Intro}

\end{document}

### 2017-03-29 chap.tex
Section 1: WIP

@@
-\documentclass{report}
+\documentclass{article}
 
@@
 \usepackage{url}
+\usepackage{natbib}
+\bibliographystyle{biblio.bst}
 
-\title{Probabilistic Forecasting of the Disturbance Storm Time Index: An Autoregressive Gaussian Process approach}
+\title{Probabilistic Forecasting of Geomagnetic Indices using Gaussian Process Models}
 
@@
 \begin{abstract}
-We present a methodology for generating probabilistic predictions for the \emph{Disturbance Storm Time} ($Dst$) geomagnetic activity index.
-We focus on the \emph{One Step Ahead} (OSA) prediction task and use the OMNI hourly resolution data to build our models.
 
-Our proposed methodology is based on the technique of \emph{Gaussian Process Regression} (GPR).
-Within this framework we develop two models; \emph{Gaussian Process Auto-Regressive} (GP-AR) and \emph{Gaussian Process Auto-Regressive with eXogenous inputs} (GP-ARX).
-
-We also propose a criterion to aid model selection with respect to the order of auto-regressive inputs.
-Finally we test the performance of the GP-AR and GP-ARX models on a set of 63 geomagnetic storms between 1998 and 2006 and illustrate sample predictions with error bars for some of these events.
+In this chapter, we give the reader an indepth view into building of probabilistic forecasting models for geomagnetic time series using the \emph{Gaussian Process} methodology outlined in the previous chapters.
 
@@
 
-\section{Intro}
+\section{Geomagnetic Time Series \& Forecasting}
+
+The Earth's magnetosphere is a region around the planet where the Earth's own magnetic field dominates the magnetic field of outer space.
+It is a region which is impinged upon constantly by the solar wind.
+The ionised plasma ejected by the sun couples with the Earth's magnetic field in a complex manner leading to highly non-linear and chaotic processes which determine the state of the magnetosphere.
+It is quite common to reduce these complex dependencies by condensing the state of the Earth's magnetosphere into a set of geomagnetic indices.
+
+Geomagnetic indices come in various forms, they may take continuous or discrete values and may be defined with varying time resolutions.
+Their values are often calculated by averaging or combining a number of readings taken by instruments, usually magnetometers, around the Earth.
+Each geomagnetic index is a proxy for a particular kind of phenomenon.
+Some popular indices are the $K_p$, $Dst$ and the $AE$ index.
+
+\begin{enumerate}
+    \item $K_p$: The Kp-index is a discrete valued global geomagnetic activity index and is based on 3 hour measurements of the K-indices \citep{Bartels}.
+The K-index itself is a three hour long quasi-logarithmic local index of the geomagnetic activity, relative to a calm day curve for the given location.
+
+    \item $AE$: The Auroral Electrojet Index, $AE$, is designed to provide a global, quantitative measure of auroral zone magnetic activity produced by enhanced Ionospheric currents flowing below and within the auroral oval \citep{AEIndex}.
+It is a continuous index which is calculated every hour.
+
+    \item $Dst$: A continuous hourly index which gives a measure of the weakening or strengthening of the Earth's equatorial magnetic field due to the weakening or strengthening of the ring currents and the geomagnetic storms \citep{DesslerAndParker}.
+\end{enumerate}
+
+Space weather forecasting systems usually use in-situ measurements of solar wind parameters, taken by satellites, as well as historical data of indices to produce forcasts for various geomagnetic time series.
+In this chapter for the purpose of exposition, we will focus on one hour ahead prediction of the $Dst$ time series.
+
+A number of modeling techniques have been applied for the prediction of the $Dst$ index.
+One of the earliest forecasting techniques involves calculating the $Dst(t)$ as a solution of an \emph{Ordinary Differential Equation} (ODE) which expressed the rate of change of $Dst(t)$ as a combination of two terms: decay and injection $\frac{d Dst(t)}{dt} = Q(t) - \frac{Dst(t)}{\tau}$, where $Q(t)$ relates to the particle injection from the plasma sheet into the inner magnetosphere.
+This method was presented first by \citet{JGR:JGR10260} and later modified and extended in works such as \citet{Wang:Dst}, \citet{JGRA:JGRA14856}, \citet{Ballatore2014} and others.
+
+%Talk about NARMAX Dst
+
+Important empirical geomagnetic prediction models include the \emph{Nonlinear Auto-Regessive Moving Average with eXogenous inputs} (NARMAX) methodology (see \citet{doi:10.1080/00207178908559767}, \citet{GRL:GRL13494}, \citet{GRL:GRL20944}, \citet{JGRA:JGRA18657}, \citet{balikhin:narmax}, \citet{JGRA:JGRA20661}, \citet{JGRA:JGRA50192}) and \emph{Artificial Neural Networks} (ANN) based models (\citet{Lund}, \citet{JGRA:JGRA17461}, \citet{SWE:SWE286}, \citet{pallocchia:hal-00318011}) for time series prediction of $Dst$ and $Kp$ indices from interplanetary magnetic field data and solar wind parameters.
+
+%Talk about need for probabilistic forecasts.
+Although much research has been done on prediction of the $Dst$ index, much less has been done on probabilistic forecasting of $Dst$.
+One such work described in \citet{McPherron:2013} involves identification of high speed solar wind streams using the WSA model, using predictions of high speed streams to construct ensembles of $Dst$ trajectories which yield the quartiles of $Dst$ time series.
+
+Probabilistic forecasting is of particular importance in geophysics applications as the end users of forecasts often require confidence bounds on the said forecasts.
+It is in this context where \emph{Gaussian Processes} become especially attractive due to their inherent probabilistic formulation and tractability of analytical inference.
+
+\section{Gaussian Processes}
+
+\emph{Gaussian Processes} first appeared in machine learning research in \citet{Neal:1996:BLN:525544}, as the limiting case of Bayesian inference performed on neural networks with infinitely many neurons in the hidden layers.
+Although their inception in the machine learning community is recent, their origins can be traced back to the geo-statistics research community where they are known as \emph{Kriging} methods \citep{krige1951statistical}.
+In pure Mathematics, \emph{Gaussian Processes} have been studied extensively and their existence was first proven by Kolmogorov's extension theorem \citep{tao2011introduction}.
+The reader is referred to \citet{Rasmussen:2005:GPM:1162254} for an in depth treatment of Gaussian Processes in machine learning.
+
+Without going into too many details, we give a quick recap of the formulation and exact inference in \emph{Gaussian Process Regression} models.
+
+\subsection{Gaussian Process Regression: Formulation}
+
+Our aim is to infer an unknown function $f(\mathbf{x})$ from its noise corrupted measurements $(\mathbf{x}_i, y_i)$ where $y_i = f(\mathbf{x}_i) + \epsilon$ and $\epsilon \sim \mathcal{N}(0, \sigma^2)$ is independent and identically distributed Gaussian noise.
+
+A \emph{Gaussian Process} model represents the finite dimensional probability distribution of $f(\mathbf{x}_i)$ by a multivariate gaussian having a particular structure for its mean and covariance as shown in \ref{eq:normal} - \ref{eq:sto}.
+
+\begin{align}
+ \mathbf{f} = & \left( \begin{array}{c} f(\mathbf{x}_1) \\ f(\mathbf{x}_2) \\ \vdots \\ f(\mathbf{x}_N) \end{array} \right) \label{eq:fvalues}\\
+ \vspace{2\baselineskip}
+ \mathbf{f} | \mathbf{x}_1, \cdots, \mathbf{x}_N \sim & \mathcal{N}\left( \mathbf{m}, \mathbf{K} \right)  \label{eq:normal}\\
+ \vspace{2\baselineskip}
+ \mathbb{P}( \mathbf{f} \ | \ \mathbf{x}_1, \cdots, \mathbf{x}_N) = & \frac{1}{(2\pi)^{n/2} det(\mathbf{K})^{1/2}} exp \left(-\frac{1}{2} (\mathbf{f} - \mathbf{m})^T \mathbf{K}^{-1} (\mathbf{f} - \mathbf{m}) \right) \label{eq:sto}
+\end{align}
+
+In order to uniquely define the distribution of $\mathbf{f}$, it is required to specify $\mathbf{m}$ and $\mathbf{K}$.
+For this probability density to be valid, there are further requirements imposed on $\mathbf{K}$:
+
+\begin{enumerate}
+      \item Symmetry: $\mathbf{K}_{ij} = \mathbf{K}_{ji} \ \forall i,j \in {1, \cdots, N} $ 
+      \item Positive Semi-definiteness: $\mathbf{z}^T \mathbf{K} \mathbf{z} \geq 0 \ \forall \mathbf{z} \in \mathbb{R}^N$  
+\end{enumerate}
+
+In \emph{Gaussian Processes} the individual elements of $\mathbf{x}$ and $\mathbf{K}$ are specified in the form of functions as shown below.
+
+\begin{align}
+      \mu_i = & \mathbb{E}[f(\mathbf{x}_i)] := m(\mathbf{x}_i) \\
+      \Lambda_{ij} = & \mathbb{E}[(f(\mathbf{x}_i) - \mu_i)(f(\mathbf{x}_j) - \mu_j)] := K(\mathbf{x}_i, \mathbf{x}_j)
+\end{align}
+
+In the machine learning community, $m(.)$ and $K(.,.)$ are known as the \emph{mean function} and \emph{covariance function} or \emph{kernel function} of the process respectively.
+Giving the expression for $m(.)$ and $K(.,.)$ uniquely specifies a particular \emph{Gaussian Process} and so a GP model is often expressed notationally as shown below.
+
+\begin{equation}
+    f(\mathbf{x}) \sim \mathcal{GP}(m(\mathbf{x}), K(\mathbf{x}, \mathbf{x}'))
+\end{equation}
+
+\subsection{Gaussian Process Regression: Inference}
+
+\bibliography{references}
 

### 2017-04-03 chap.tex
Working on model selection section

@@
 
-Without going into too many details, we give a quick recap of the formulation and exact inference in \emph{Gaussian Process Regression} models.
+Without going into too many details, we give a quick recap of the formulation and exact inference in \emph{Gaussian Process Regression} (GPR) models.
 
@@
 In the machine learning community, $m(.)$ and $K(.,.)$ are known as the \emph{mean function} and \emph{covariance function} or \emph{kernel function} of the process respectively.
-Giving the expression for $m(.)$ and $K(.,.)$ uniquely specifies a particular \emph{Gaussian Process} and so a GP model is often expressed notationally as shown below.
+Giving a closed form expression for $m(.)$ and $K(.,.)$ uniquely specifies a particular \emph{Gaussian Process} and so a GP model is often expressed notationally as shown below.
 
@@
 
+In order to generate predictions $f(\mathbf{x}^{*}_i)$ for a set of test points $ {\mathbf{x}^{*}_i : \forall i \in 1, \cdots, M} $.
+Using the multivariate Gaussian distribution in equation (\ref{eq:sto}) we can construct the joint distribution of $f(\mathbf{x})$ over the training and test points.
+
+\begin{align}
+    \mathbf{f}_* = & \left( \begin{array}{c} f(\mathbf{x^{*}_1}) \\ f(\mathbf{x^{*}_2}) \\ \vdots \\ f(\mathbf{x^{*}_M}) \end{array} \right)_{M \times 1} \\
+     \vspace{4\baselineskip}
+    \left( \begin{array}{c} \mathbf{y} \\ \mathbf{f_*} \end{array} \right) | \ \ \mathbf{X}, \mathbf{X}_* \sim & 
+    \mathcal{N}\left(\mathbf{0}, \left[ \begin{array}{cc} \mathbf{K} + \sigma^{2} \mathbf{I} & \mathbf{K}_{*} \\ \mathbf{K}_{*}^T & \mathbf{K}_{**} \end{array} \right ] \right) \label{eq:dist}
+\end{align}
+
+Probabilistic predictions $f_*$ can be generated by constructing the conditional distribution $\mathbf{f_*}|\mathbf{X},\mathbf{y},\mathbf{X_*}$ which is also a multivariate gaussian as shown in equation \ref{eq:posterior}.
+
+\begin{align}
+    & \mathbf{f_*}|\mathbf{X},\mathbf{y},\mathbf{X_*} \sim \mathcal{N}(\mathbf{\bar{f}_*}, \Sigma_*)  \label{eq:posterior} \\
+    & \mathbf{\bar{f}_*} =  \ \mathbf{K}^T_{*} [\mathbf{K} + \sigma^{2} \mathbf{I}]^{-1} \mathbf{y} \label{eq:posteriormean} \\
+    & \Sigma_* = \ \mathbf{K}_{**} - \mathbf{K}^T_{*} \left(\mathbf{K} + \sigma^{2} \mathbf{I}\right)^{-1} \mathbf{K}_{*} \label{eq:posteriorcov}
+\end{align}
+
+\section{One Hour Ahead $Dst$: Formulation} \label{sec:osa}
+
+Below in equations (\ref{eq:Dst}) - (\ref{eq:DstGP}) we outline a \emph{Gaussian Process} formulation for \emph{OSA} prediction of $Dst$.
+A vector of features $\mathbf{x}_{t-1}$ is used as input to an unknown function $f(\mathbf{x}_{t-1})$.
+
+The features $\mathbf{x}_{t-1}$ can be any collection of quantities in the hourly resolution OMNI data set.
+Generally $\mathbf{x}_{t-1}$ are time histories of $Dst$ and other important variables such as plasma pressure $p(t)$, solar wind speed $V(t)$, $z$ component of the interplanetary magnetic field $B_z(t)$.
+
+\begin{align}
+    Dst(t) & =  f(\mathbf{x}_{t-1}) + \epsilon \label{eq:Dst} \\
+    \epsilon & \sim  \mathcal{N}(0, \sigma^2) \label{eq:GPNoise} \\
+    f(x_t) & \sim  \mathcal{GP}(m(\mathbf{x}_t), K_{osa}(\mathbf{x}_t, \mathbf{x}_s)) \label{eq:DstGP} \\
+\end{align}
+
+We consider two choices for the input features $\mathbf{x}_{t-1}$ leading to two variants of \emph{Gaussian Process} regression for $Dst$ time series prediction.
+
+\subsection{Gaussian Process Auto-Regressive (GP-AR)} \label{sec:gpar}
+
+The simplest auto-regressive models for \emph{OSA} prediction of $Dst$ are those that use only the history of $Dst$ to construct input features for model training.
+The input features $\mathbf{x}_{t-1}$ at each time step are the history of $Dst(t)$ until a time lag of $p$ hours.
+
+\begin{align*}
+    \mathbf{x}_{t-1} = & \left(Dst(t-1), \cdots , Dst(t-p+1)\right)
+\end{align*}
+
+\subsection{Gaussian Process Auto-Regressive with eXogenous inputs (GP-ARX)} \label{sec:gparx}
+
+Auto-regressive models can be augmented by including exogenous quantities in the inputs $\mathbf{x}_{t-1}$ at each time step, in order to improve predictive accuracy.
+$Dst$ gives a measure of ring currents, which are modulated by plasma sheet particle injections into the inner magnetosphere during sub-storms.
+Studies have shown that the substorm occurrence rate increases with solar wind velocity (high speed streams) \citet{Kissinger2011,Newell2016}.
+Prolonged southward interplanetary magnetic field (IMF) $z$-component ($B_z$) is needed for sub-storms to occur \citet{McPherron1986}.
+An increase in the solar wind electric field, $V_{sw}B_z$, can increase the dawn-dusk electric field in the magnetotail, which in turn determines the amount of plasma sheet particle that move to the inner magnetosphere \citet{Friedel2001}.
+Therefore, our exogenous parameters consist of solar wind velocity $V_{sw}$ and IMF $B_z$.
+
+In this model we choose distinct time lags $p$, $p_{v}$ and $p_{b}$ for $Dst$, $V$ and $B_z$ respectively.
+
+\begin{align*}
+       \mathbf{x}_{t-1} & = (Dst(t-1), \cdots , Dst(t-p+1), \\
+        & \ \ \ \ \  V_{sw}(t-1), \cdots, V_{sw}(t-p_{v}+1),\\
+        & \ \ \ \ \  B_{z}(t-1), \cdots, B_{z}(t-p_{b}+1))
+\end{align*}
+
+\section{One Hour Ahead $Dst$: Model Design}
+
+\subsection{Choice of Mean Function}
+
+Mean functions in GPR models encode trends in the data, they are the baseline predictions the model falls back to in case the training and test data have little correlation as predicted by the kernel function.
+If there is no prior knowledge about the function to be approximated, \citet{Rasmussen:2005:GPM:1162254} state that it is perfectly reasonable to choose $m(\mathbf{x} = 0)$ as the mean function, as long as the target values are normalized.
+In the case of the $Dst$ time series, it is known that the so called \emph{persistence model} $\hat{D}st(t) = Dst(t-1)$ performs quite well in the context of OSA prediction.
+We therefore choose the \emph{persistence model} as the mean function in our OSA Dst models.
+
+\subsection{Choice of Kernel}
+
+For the success of a \emph{Gaussian Process} model an appropriate choice of kernel function is paramount.
+The symmetry and positive semi-definiteness of \emph{Gaussian Process} kernels implies that they represent inner-products between some basis function representation of the data.
+The interested reader is suggested to refer to \citet{Berlinet2004}, \citet{Scholkopf:2001:LKS:559923} and \citet{hofmann2008} for a thorough treatment of kernel functions and the rich theory behind them.
+Some common kernel functions used in machine learning are listed in Table \ref{table:kernel}.
+
+\begin{table}[h]
+\caption{Popular Kernel functions used in GPR models}
+\centering
+\begin{tabular}{l c c}
+\hline
+ Name  & Expression & Hyperparameters  \\
+\hline
+  Radial Basis Function (RBF)  & $\frac{1}{2} exp(-||\mathbf{x} - \mathbf{y}||^2/l^2)$  & $l \in \mathbb{R}$   \\
+
+  Polynomial  & $(\mathbf{x}^\intercal \mathbf{y} + b)^d$ & $b \in \mathbb{R}, d \in \mathbb{N}$   \\
+
+  Laplacian  & $exp(-||\mathbf{x} - \mathbf{y}||_{1}/\theta)$  & $\theta \in \mathbb{R}^+$  \\
+
+  Student's T  & $1/(1 + ||\mathbf{x} - \mathbf{y}||_{2}^d)$ & $d \in \mathbb{R}^{+}$\\
+
+  Maximum Likelihood Perceptron  & $sin^{-1}(\frac{w\mathbf{x}^\intercal \mathbf{y} + b}{\sqrt{w\mathbf{x}^\intercal \mathbf{x} + b + 1} \sqrt{w\mathbf{y}^\intercal \mathbf{y} + b + 1}})$ & $w, b \in \mathbb{R}^{+}$\\
+\hline
+\end{tabular}
+\label{table:kernel}
+\end{table}
+
+The quantities $l$ in the RBF, and $b$ and $d$ in the polynomial kernel are known as \emph{hyper-parameters}.
+Hyper-parameters give flexibility to a particular kernel structure, for example $d = 1, 2, 3, \cdots$ in the polynomial kernel represents linear, quadratic, cubic and higher order polynomials respectively.
+The process of assigning values to the \emph{hyper-parameters} is crucial in the model building process and is known as \emph{model selection}.
+
+In this text, we construct GPR models with a combination of the \emph{maximum likelihood perceptron} kernel and \emph{student's T} kernel as shown in equations (\ref{eq:usedKernel}).
+The \emph{maximum likelihood perceptron} kernel is the \emph{Gaussian Process} equivalent of a single hidden layer feed-forward neural network model as demonstrated in \citet{Neal:1996:BLN:525544}.
+
+\begin{align}
+    K_{osa}(\mathbf{x}, \mathbf{y}) & = K_{mlp}(\mathbf{x}, \mathbf{y}) + K_{st}(\mathbf{x}, \mathbf{y}) \label{eq:usedKernel} \\
+    K_{mlp}(\mathbf{x}, \mathbf{y}) & = sin^{-1}(\frac{w\mathbf{x}^\intercal \mathbf{y} + b}{\sqrt{w\mathbf{x}^\intercal \mathbf{x} + b + 1} \sqrt{w\mathbf{y}^\intercal \mathbf{y} + b + 1}}) \\
+    K_{st}(\mathbf{x}, \mathbf{y}) & = \frac{1}{1 + ||\mathbf{x} - \mathbf{y}||_{2}^d}
+\end{align}
+
+\subsection{Model Selection}
+
+Given a GPR model with a kernel function $K_\theta$, the problem of model selection consists of finding appropriate values for the kernel hyper-parameters $\theta = \left(\theta_1, \theta_2, \cdots, \theta_i\right)$.
+In order to assign a value to $\theta$, we must define an objective function which represents our confidence that the GP model built from a particular value of $\theta$ is the best performing model.
+Since GP models are constructed from assumptions about the conditional probability distribution of the data $\mathbb{P}(\mathbf{y}|\mathbf{X})$, it is natural to use the negative log-likelihood of the training data as a model selection criterion.
+
+\begin{align*}
+  Q(\theta) & = - log(p(\mathbf{y}|\mathbf{X}, K_\theta)) \\
+            & = -\frac{1}{2} \mathbf{y}^\intercal (\mathbf{K}_\theta + \sigma^{2} \mathbf{I})^{-1} \mathbf{y} - \frac{1}{2}|\mathbf{K}_\theta + \sigma^{2} \mathbf{I}| - \frac{N}{2}log(2\pi) \\
+  \mathbf{K}_\theta & = [K_{\theta}(\mathbf{x}_i, \mathbf{x}_j)]_{N \times N}
+\end{align*}
+
+The model selection problem can now be expressed as the minimization problem shown below.
+
+\begin{align*}
+\theta^* = arg\min_{\theta} \ Q(\theta)
+\end{align*}
+
+The objective function $Q(\theta)$ can have multiple local minima, and evaluating the value of $Q(.)$ at any given $\theta$ requires inversion of the matrix $\mathbf{K}_\theta + \sigma^{2} \mathbf{I}$ which has a time complexity $O(N^3)$ as noted above.
+In the interest of saving computational cost, one cannot use exhaustive search through the domain of the hyper-parameters to inform our choice for $\theta$.
+
+The model selection problem has historically received less attention due to a combination of difficulty and numerous design decisions like choice of hyper-parameter optimization techniques, which makes model selection as much of an art as science.
+
+Choosing a suitable heuristic or algorithm for model selection for GP models is quite often a balancing exercise between the computational limitations of the modeler and the need to explore the hyper-parameter space in a manner that yields a satisfactory predictive model.
+Some of the techniques used for model selection in the context of GPR include.
+
+\subsubsection*{Grid Search}
+
+This is the simplest procedure that may be applied for selection of GPR models.
+This routine constructs a grid of values for $\theta$ as the cartesian product of one dimensional grids for each $\theta_i$, evaluate $Q(.)$ at each such grid point and choose the configuration which yields minimum value of $Q(.)$.
+The advantage of this technique is its simplicity and control over the computational cost of training GP models for practical problems.
+The computational cost of this procedure scales with the total number of points on the grid, this is controlled by the scales and number of grid points for each hyper-parameter $\theta_i$.
+
+\subsubsection*{Coupled Simulated Annealing}
+
+Introduced in \citet{Xavier-De-Souza2010}, it follows the same procedure as \emph{grid search}, but after evaluation of the objective $Q(.)$ on the grid, each grid point is iteratively mutated in a random walk fashion.
+This mutation is accepted or rejected according to the new value of $Q(.)$ as well as its value on the other grid points.
+This procedure is iterated until some stop criterion is reached.
+
+\subsubsection*{Maximum Likelihood}
+
+This technique as outlined in \citet{Rasmussen:2005:GPM:1162254} is a form of \emph{gradient descent}.
+It involves starting with an initial guess for $\theta$ and iteratively improving it by calculating the gradient of $Q(.)$ with respect to $\theta$.
+Although this method seems intuitive, it introduces an extra computational cost of calculating the gradient of $Q(\theta)$ with respect to each $\theta_i$ in every iteration and applying this method can sometimes lead to overfitting of the GPR model to the training data \citep{Rasmussen:2005:GPM:1162254}.
+
 \bibliography{references}

### 2017-04-07 chap.tex
Added ML-II section

@@
 In order to assign a value to $\theta$, we must define an objective function which represents our confidence that the GP model built from a particular value of $\theta$ is the best performing model.
-Since GP models are constructed from assumptions about the conditional probability distribution of the data $\mathbb{P}(\mathbf{y}|\mathbf{X})$, it is natural to use the negative log-likelihood of the training data as a model selection criterion.
+Since GP models are constructed from assumptions about the conditional probability distribution of the data $p(\mathbf{y}|\mathbf{X})$, it is natural to use the negative log-likelihood of the training data as a model selection criterion.
 
 \begin{align*}
-  Q(\theta) & = - log(p(\mathbf{y}|\mathbf{X}, K_\theta)) \\
-            & = -\frac{1}{2} \mathbf{y}^\intercal (\mathbf{K}_\theta + \sigma^{2} \mathbf{I})^{-1} \mathbf{y} - \frac{1}{2}|\mathbf{K}_\theta + \sigma^{2} \mathbf{I}| - \frac{N}{2}log(2\pi) \\
+  Q(\theta) & = - log \ p(\mathbf{y}|\mathbf{X}, \theta) \\
+            & = \frac{1}{2} \mathbf{y}^\intercal (\mathbf{K}_\theta + \sigma^{2} \mathbf{I})^{-1} \mathbf{y} + \frac{1}{2}|\mathbf{K}_\theta + \sigma^{2} \mathbf{I}| + \frac{N}{2}log(2\pi) \\
   \mathbf{K}_\theta & = [K_{\theta}(\mathbf{x}_i, \mathbf{x}_j)]_{N \times N}
@@
 
-Introduced in \citet{Xavier-De-Souza2010}, it follows the same procedure as \emph{grid search}, but after evaluation of the objective $Q(.)$ on the grid, each grid point is iteratively mutated in a random walk fashion.
+Introduced in \citet{Xavier-De-Souza2010}, \emph{Coupled Simulated Annealing} (CSA) is a generalisation of \emph{Simulated Annealing} (SA).
+
+CSA follows the same procedure as \emph{grid search}, but after evaluation of the objective $Q(.)$ on the grid, each grid point is iteratively mutated in a random walk fashion.
 This mutation is accepted or rejected according to the new value of $Q(.)$ as well as its value on the other grid points.
@@
 
-This technique as outlined in \citet{Rasmussen:2005:GPM:1162254} is a form of \emph{gradient descent}.
+This technique, as described in \citet{Rasmussen:2005:GPM:1162254}, is a form of \emph{gradient descent}.
 It involves starting with an initial guess for $\theta$ and iteratively improving it by calculating the gradient of $Q(.)$ with respect to $\theta$.
-Although this method seems intuitive, it introduces an extra computational cost of calculating the gradient of $Q(\theta)$ with respect to each $\theta_i$ in every iteration and applying this method can sometimes lead to overfitting of the GPR model to the training data \citep{Rasmussen:2005:GPM:1162254}.
+The gradient of $Q(.)$ with respect to each hyper-parameter $\theta_i$ can be calculated analytically as shown below.
+
+\begin{align*}
+\frac{\partial}{\partial \theta_i} log \ p(\mathbf{y}|\mathbf{X}, \theta) & = \frac{1}{2} \mathbf{y}^\intercal K^{-1} \frac{\partial K}{\partial \theta_i} K^{-1} \mathbf{y}  - \frac{1}{2} tr(K^{-1} \frac{\partial K}{\partial \theta_i})\\
+ & = \frac{1}{2} tr((\mathbf{\alpha}\mathbf{\alpha}^\intercal - K^{-1}) \frac{\partial K}{\partial \theta_i}) \ \ where \ \mathbf{\alpha} = K^{-1} \mathbf{y}
+\end{align*}
+
+Although gradient based \emph{Maximum Likelihood} looks very similar to the \emph{Gradient Descent} used to learn parametric regression models like \emph{Feed forward Neural Networks}, in the GPR model selection context it introduces an extra computational cost of calculating the gradient of $Q(\theta)$ with respect to each $\theta_i$ in every iteration.
+Applying this method can sometimes lead to overfitting of the GPR model to the training data \citep{Rasmussen:2005:GPM:1162254}.
 

### 2017-04-07 chap.tex
CSA section [WIP]

@@
 
-Introduced in \citet{Xavier-De-Souza2010}, \emph{Coupled Simulated Annealing} (CSA) is a generalisation of \emph{Simulated Annealing} (SA).
+Introduced in \citet{Xavier-De-Souza2010}, \emph{Coupled Simulated Annealing} (CSA) is a family of optimization techniques which is a generalization of \emph{Simulated Annealing} (SA) \citep{Kirkpatrick671}.
+The CSA optimization technique can be understood as a set of parallel SA processes which are coupled by their acceptance probabilities.
 
-CSA follows the same procedure as \emph{grid search}, but after evaluation of the objective $Q(.)$ on the grid, each grid point is iteratively mutated in a random walk fashion.
+From an implementation point of view, CSA follows the same procedure as \emph{grid search}, but after evaluation of the objective $Q(.)$ on the grid, each grid point is iteratively mutated in a random walk fashion.
 This mutation is accepted or rejected according to the new value of $Q(.)$ as well as its value on the other grid points.

### 2017-04-07 chap.tex
CSA: Section Complete

@@
 
+To illustrate how CSA updates are made and accepted consider a population or ensemble of hyper-parameter values ${x_i}$ which can be considered to be initialised in a grid based fashion.
+For each configuration $x_i$ a probing solution $y_i = x_i + \epsilon$ is generated where $\epsilon$ is drawn from some random variable.
+The acceptance probability of the solution is calculated as shown below.
+
+\begin{align*}
+\mathbb{P}(x_i \rightarrow y_i | {x_i}) & = \frac{exp(-E(y_i)/T)}{exp(-E(y_i)/T) + \gamma} \\
+\gamma & = \sum_{j}{exp \left ( \frac{-E(x_j)}{T} \right )}
+\end{align*}
+
+The quantity $\gamma$ defines the \emph{coupling} between the states ${x_i}$ of the ensemble, when the number of configurations in the emsemble $x_i$ is one this rule reduces to the classical \emph{Metropolis-Hastings} acceptance rule of the SA algorithm.
+Upto 4 variants of the CSA algorithm have been defined with different acceptance rules and temperature annealing schedules, the reader is advised to refer to \citet{Xavier-De-Souza2010} for an in-depth treatment of the CSA optimization class.
+
 \subsubsection*{Maximum Likelihood}

### 2017-04-09 chap.tex
[WIP]: Model order selection et. al

@@
 
-\subsection{Model Selection}
+\subsection{Model Selection: Hyper-parameters}
 
@@
 
+\citet{Rasmussen:2005:GPM:1162254} note that there are ready interpretations to the three terms which add up to give $Q(.)$.
+The first term $\frac{1}{2} \mathbf{y}^\intercal (\mathbf{K}_\theta + \sigma^{2} \mathbf{I})^{-1} \mathbf{y}$ quantifies the fit of the model with respect to the training data, the second term $\frac{1}{2}|\mathbf{K}_\theta + \sigma^{2} \mathbf{I}|$ is a complexity penalty, in non-parametric models such as \emph{Gaussian Processes} the model complexity grows with the number of training data points $N$.
+The third term $\frac{N}{2}log(2\pi)$ is a normalization term.
+
 The objective function $Q(\theta)$ can have multiple local minima, and evaluating the value of $Q(.)$ at any given $\theta$ requires inversion of the matrix $\mathbf{K}_\theta + \sigma^{2} \mathbf{I}$ which has a time complexity $O(N^3)$ as noted above.
@@
 The model selection problem has historically received less attention due to a combination of difficulty and numerous design decisions like choice of hyper-parameter optimization techniques, which makes model selection as much of an art as science.
-
 Choosing a suitable heuristic or algorithm for model selection for GP models is quite often a balancing exercise between the computational limitations of the modeler and the need to explore the hyper-parameter space in a manner that yields a satisfactory predictive model.
@@
 
-The quantity $\gamma$ defines the \emph{coupling} between the states ${x_i}$ of the ensemble, when the number of configurations in the emsemble $x_i$ is one this rule reduces to the classical \emph{Metropolis-Hastings} acceptance rule of the SA algorithm.
-Upto 4 variants of the CSA algorithm have been defined with different acceptance rules and temperature annealing schedules, the reader is advised to refer to \citet{Xavier-De-Souza2010} for an in-depth treatment of the CSA optimization class.
+The quantity $\gamma$ defines the \emph{coupling} between the states ${x_i}$ of the ensemble, when the number of configurations in the ensemble $x_i$ is one this rule reduces to the classical \emph{Metropolis-Hastings} acceptance rule of the SA algorithm.
+Up to 4 variants of the CSA algorithm have been defined with different acceptance rules and temperature annealing schedules, the reader is advised to refer to \citet{Xavier-De-Souza2010} for an in-depth treatment of the CSA optimization class.
 
@@
 
+\subsection{Model Selection: Auto-regressive Order}
+
+Apart from having a practical method for choosing values of kernel hyper-parameters, we must also address the issue of choosing the time delays $p$ and $p_v, p_b$ for the GP-AR and GP-ARX models.
+One might argue that we group these quantities with the kernel hyper-parameters in the model selection procedure, in the context of the GP-AR and GP-ARX models there is one reason why this might not be beneficial.
+
+The value of kernel hyper-parameters $\theta^*$ which maximise the marginal likelihood $Q(.)$ are conditional on the chosen model orders $p$ and $p, p_v, p_b$ for GP-AR and GP-ARX respectively.
+Since the model orders determine the size of the input space of the predictive models, they influence the values of characteristic length scales and other kernel hyper-parameters which yield a good fit.
+
+Not only do the model orders influence the value of kernel hyper-parameters $\theta^*$ which maximise model fit, but increasing values of $p, p_v, p_b$ also make it more difficult for the model selection algorithms to search the hyper-parameter space, due to the curse of dimensionality.
+
+An empirical way to choose the values of $p, p_v, p_b$ for the GP-AR and GP-ARX models is by choosing an independent validation data set which is used to evaluate the GP models obtained from the model selection procedures of the previous section.
+It is then possible for the modeler to choose these model orders in way to balance performance and dimensionality of input features.
+
 \bibliography{references}

### 2017-04-10 chap.tex
Appended abstract

@@
 In this chapter, we give the reader an indepth view into building of probabilistic forecasting models for geomagnetic time series using the \emph{Gaussian Process} methodology outlined in the previous chapters.
+We highlight design decisions and practical issues must be addressed in order to build \emph{Gaussian Process} models for probabilistic prediction of a quantity of interest.
+As a case study, we formulate predictive models for one hour ahead prediction of the $Dst$ geomagnetic index.
 
@@
 
-This is the simplest procedure that may be applied for selection of GPR models.
+This is the simplest procedure that may be applied for performing model selection in GPR models.
 This routine constructs a grid of values for $\theta$ as the cartesian product of one dimensional grids for each $\theta_i$, evaluate $Q(.)$ at each such grid point and choose the configuration which yields minimum value of $Q(.)$.

### 2017-04-10 chap.tex
Added sections on OMNI data

@@
 Space weather forecasting systems usually use in-situ measurements of solar wind parameters, taken by satellites, as well as historical data of indices to produce forcasts for various geomagnetic time series.
+The principal data source for forecasting of geomagnetic indices is the \emph{OMNI} data set hosted by NASA on the \emph{Space Physics Data Facility} (SPDF).
+
+\section{Data Source: OMNI}
+The \emph{OMNI} data set is a collation of in-situ measurements taken by orbiting satellites and ground based measurements of geomagnetic indices.
+The data itself is represented as time averaged versions of the various quantities and is available at two frequencies, by hour and by minute.
+The quantities recorded can be broadly classified into two groups.
+
+\begin{enumerate}
+
+  \item Plasma Parameters and Magnetic fields: Measurements of the solar wind particularly the solar wind speed, flow angles, pressure, proton temperature, proton density and field measurements such as strength of electric and magnetic fields as well as their components in various geomagnetic coordinate systems.
+
+  \item Indices: $Kp$, $Ae$, $Ap$, $Dst$, sunspot number.
+
+\end{enumerate}
+
+To get a more indepth understanding about the \emph{OMNI} data, interested readers are encouraged to visit the NASA omni web service (\url{https://omniweb.gsfc.nasa.gov}).
+
 In this chapter for the purpose of exposition, we will focus on one hour ahead prediction of the $Dst$ time series.
+From the data we will use time histories of $Dst$, solar wind speed $V$ and the $z$-component of the interplanetary magnetic field $B_z$.
+Our choice is motivated by both physical and practical considerations: solar wind speed and interplanetary magnetic field are important external drivers for particle injection into the magnetosphere and they have large coverage in the \emph{OMNI} data with relatively lesser number of missing records.
+
+\section{$Dst$ Forecasting}
+
+\subsection{State of the art}
 
@@
 %Talk about NARMAX Dst
-
 Important empirical geomagnetic prediction models include the \emph{Nonlinear Auto-Regessive Moving Average with eXogenous inputs} (NARMAX) methodology (see \citet{doi:10.1080/00207178908559767}, \citet{GRL:GRL13494}, \citet{GRL:GRL20944}, \citet{JGRA:JGRA18657}, \citet{balikhin:narmax}, \citet{JGRA:JGRA20661}, \citet{JGRA:JGRA50192}) and \emph{Artificial Neural Networks} (ANN) based models (\citet{Lund}, \citet{JGRA:JGRA17461}, \citet{SWE:SWE286}, \citet{pallocchia:hal-00318011}) for time series prediction of $Dst$ and $Kp$ indices from interplanetary magnetic field data and solar wind parameters.
 
+\subsection{Probabilistic Forecasting}
 %Talk about need for probabilistic forecasts.

### 2017-04-10 chap.tex
Added method summary

@@
     \item $Dst$: A continuous hourly index which gives a measure of the weakening or strengthening of the Earth's equatorial magnetic field due to the weakening or strengthening of the ring currents and the geomagnetic storms \citep{DesslerAndParker}.
+A large negative value of $Dst$ ($ \leq -150 \ nT$) indicates occurence of geomagnetic storm events which are of particular importance in space weather prediction due to their effects on telecommunications infrastructure.
 \end{enumerate}
@@
 
-\section{One Hour Ahead $Dst$: Model Design}
+\section{One Hour Ahead $Dst$: Model Design}\label{sec:modeldesign}
 
@@
 
+\section{GP-AR and GP-ARX: Workflow Summary}
+
+Section \ref{sec:modeldesign} brings together all the practical issues with respect to building of GP-AR and GP-ARX models for predicting $Dst$, they can be condensed to form the following steps.
+
+\begin{enumerate}
+
+\item \textbf{Data Preprocessing}: Three seperate data sets for model training and hyper-parameter selection, model order selection and model evaluation must be created.
+Care must be taken that the size of the training sets does not delay the processes of kernel hyper-parameter selection.
+The training, validation and test data sets created from this procedure must all contain non-overlapping geomagnetic storm events.
+
+\item \textbf{Choice of model components}:  Kernel, mean function, model selection procedure.
+
+\item \textbf{Model Order Selection}: In order to find appropriate values for the auto-regressive time lags $p, p_v, p_b$ for $Dst$, solar wind speed and IMF $B_z$ respectively, create models upto a certain maximum time lag i.e.
+$3 \leq p \leq 8$ for GP-AR and $3 \leq p + p_v + p_b \leq 8$ for GP-ARX.
+For each such model perform the model selection routine to select values of the kernel hyper-parameters and evaluate the resultant model on the validation set.
+Choose the value of $p, p_v, p_b$ which maximises predictive performance on the validation set.
+
+\end{enumerate}
+
 \bibliography{references}

### 2017-04-18 chap.tex
Completing Workflow section

@@
 \usepackage{natbib}
+\usepackage{rotating}
+\usepackage{adjustbox}
+\graphicspath{{figures/}}
 \bibliographystyle{biblio.bst}
@@
 
+We selected OMNI data sections 00:00 January 3 2010 to 23:00 January 23 2010 and 20:00 August 5 2011 to 22:00 August 6 2011 for training the GP-AR and GP-ARX models.
+The first training data section consists of ambient fluctuations of $Dst$ while the second contains a geomagnetic storm.
+A set of 24 storm events listed in Table \ref{table:validationstorms} is kept aside as a validation data set for selecting model orders $p, p_v, p_b$ while a list of 63 storm events listed in table \ref{table:teststorms} is used for final model evaluation.
+
 \item \textbf{Choice of model components}:  Kernel, mean function, model selection procedure.
+
+We choose a kernel $K_{osa}$ of the form given in equation \ref{eq:usedKernel}.
+In order to find appropriate values of the hyper-parameters of $K_{osa}$, we apply \emph{grid search}, \emph{coupled simulated annealing} and \emph{maximum likelihood} methods.
+We fix the parameters $d$ and $\sigma^2$ of $K_{st}$ and model noise to values $0.01$ and $0.2$ respectively, the remaining parameters $w$ and $b$ are kept free to be calculated by model selection.
+Table \ref{table:modelselection} summarizes the settings used to run each model selection procedure.
+
+\item \textbf{Choice of Performance Metrics}: To compare the performance of GP and GP-ARX models of varying auto-regressive orders, their predictive performance during the validation (and later final evaluation) phase must be calculated using some clearly defined metrics.
+We use the following as performance metrics for our modeling experiments.
+
+\begin{enumerate}
+    \item The mean absolute error.
+    \begin{equation}
+        MAE = \sum_{t=1}^{n} \left |(Dst(t) - \hat{D}st(t)) \right | / n
+    \end{equation}
+    \item The root mean square error.
+    \begin{equation}
+        RMSE = \sqrt{\sum_{t=1}^{n} (Dst(t) - \hat{D}st(t))^2 / n}
+    \end{equation}
+    \item Correlation coefficient between the predicted and actual value of $Dst$.
+    \begin{equation}
+        CC = Cov(Dst, \hat{D}st)/\sqrt{Var(Dst) Var(\hat{D}st)}
+    \end{equation}
+\end{enumerate}
 
@@
 
+\item \textbf{Final Evaluation}: After selecting GP-AR and GP-ARX models with particular values of auto-regressive orders and kernel hyper-parameters in the model selection phase, the best performing GP-AR and GP-ARX models are then evaluated on the test data set in order to get an unbiased picture of their predictive performance.
+In our case we use the already chosen test set outlined in table \ref{table:teststorms} for final model evaluation.
+The GP based OSA prediction models are compared against the naive \emph{Persistence} model which serves as a performance baseline for the OSA prediction of $Dst$.
+
 \end{enumerate}
+
+\section{Experiments \& Results}
+
+\begin{table}[h]
+\centering
+\caption{Settings of model selection procedures}
+\begin{tabular}{l c c c}
+\hline
+Procedure & Grid Size & Step & Max Iterations \\
+\hline
+Grid Search & 10 & 0.2 & NA \\
+Coupled Simulated Annealing & 4 & 0.2 & 30 \\
+Maximum likelihood & NA & 0.2 & 150\\
+\end{tabular}
+\label{table:modelselection}
+\end{table}
+
+\begin{table}[h]
+\centering
+\caption{Storm events used for model selection of GP-AR and GP-ARX}
+\label{table:validationstorms}
+\begin{tabular}{llllll}
+\hline
+Event Id & Start Date & Start Hour & End Date & End Hour & min. Dst \\ \hline
+1 & 1995/03/26 & 0500 & 1995/03/26 & 2300 & −107 \\
+2 & 1995/04/07 & 1300 & 1995/04/09 & 0900 & −149 \\
+3 & 1995/09/27 & 0100 & 1995/09/28 & 0400 & −108 \\
+4 & 1995/10/18 & 1300 & 1995/10/19 & 1400 & −127 \\
+5 & 1996/10/22 & 2200 & 1996/10/23 & 1100 & −105 \\
+6 & 1997/04/21 & 1000 & 1997/04/22 & 0900 & −107 \\
+7 & 1997/05/15 & 0300 & 1997/05/16 & 0000 & −115 \\
+8 & 1997/10/10 & 1800 & 1997/10/11 & 1900 & −130 \\
+9 & 1997/11/07 & 0000 & 1997/11/07 & 1800 & −110 \\
+10 & 1997/11/22 & 2100 & 1997/11/24 & 0400 & −108 \\
+11 & 2005/06/12 & 1700 & 2005/06/13 & 1900 & −106 \\
+12 & 2005/08/31 & 1200 & 2005/09/01 & 1200 & −122 \\
+13 & 2006/12/14 & 2100 & 2006/12/16 & 0300 & −162 \\
+14 & 2011/09/26 & 1400 & 2011/09/27 & 1200 & −101 \\
+15 & 2011/10/24 & 2000 & 2011/10/25 & 1400 & −132 \\
+16 & 2012/03/08 & 1200 & 2012/03/10 & 1600 & −131 \\
+17 & 2012/04/23 & 1100 & 2012/04/24 & 1300 & −108 \\
+18 & 2012/07/15 & 0100 & 2012/07/16 & 2300 & −127 \\
+19 & 2012/09/30 & 1300 & 2012/10/01 & 1800 & −119 \\
+20 & 2012/10/08 & 0200 & 2012/10/09 & 1700 & −105 \\
+21 & 2012/11/13 & 1800 & 2012/11/14 & 1800 & −108 \\
+22 & 2013/03/17 & 0700 & 2013/03/18 & 1000 & −132 \\
+23 & 2013/05/31 & 1800 & 2013/06/01 & 2000 & −119 \\
+24 & 2014/02/18 & 1500 & 2014/02/19 & 1600 & −112
+\end{tabular}
+\end{table}
+
+\begin{table}[h]
+\fontsize{6}{7.5}\selectfont
+\centering
+\caption{Storm events used to evaluate GP-AR and GP-ARX models}
+\label{table:teststorms}
+\begin{tabular}{cccccc}
+\hline
+Event Id & Start Date & Start Time & End Date & End Time & min. Dst \\ \hline
+1 & 1998/02/17 & 1200 & 1998/02/18 & 1000 & -100 \\
+2 & 1998/03/10 & 1100 & 1998/03/11 & 1800 & -116 \\
+3 & 1998/05/04 & 0200 & 1998/05/05 & 0200 & -205 \\
+4 & 1998/08/26 & 1000 & 1998/08/29 & 0700 & -155 \\
+5 & 1998/09/25 & 0100 & 1998/09/26 & 0000 & -207 \\
+6 & 1998/10/19 & 0500 & 1998/10/20 & 0800 & -112 \\
+7 & 1998/11/09 & 0300 & 1998/11/10 & 1600 & -142 \\
+8 & 1998/11/13 & 0000 & 1998/11/15 & 0400 & -131 \\
+9 & 1999/01/13 & 1600 & 1999/01/14 & 2000 & -112 \\
+10 & 1999/02/18 & 0300 & 1999/02/19 & 2100 & -123 \\
+11 & 1999/09/22 & 2000 & 1999/09/23 & 2300 & -173 \\
+12 & 1999/10/22 & 0000 & 1999/10/23 & 1400 & -237 \\
+13 & 2000/02/12 & 0500 & 2000/02/13 & 1500 & -133 \\
+14 & 2000/04/06 & 1700 & 2000/04/08 & 0900 & -288 \\
+15 & 2000/05/24 & 0100 & 2000/05/25 & 2000 & -147 \\
+16 & 2000/08/10 & 2000 & 2000/08/11 & 1800 & -106 \\
+17 & 2000/08/12 & 0200 & 2000/08/13 & 1700 & -235 \\
+18 & 2000/10/13 & 0200 & 2000/10/14 & 2300 & -107 \\
+19 & 2000/10/28 & 2000 & 2000/10/29 & 2000 & -127 \\
+20 & 2000/11/06 & 1300 & 2000/11/07 & 1800 & -159 \\
+21 & 2000/11/28 & 1800 & 2000/11/29 & 2300 & -119 \\
+22 & 2001/03/19 & 1500 & 2001/03/21 & 2300 & -149 \\
+23 & 2001/03/31 & 0400 & 2001/04/01 & 2100 & -387 \\
+24 & 2001/04/11 & 1600 & 2001/04/13 & 0700 & -271 \\
+25 & 2001/04/18 & 0100 & 2001/04/18 & 1300 & -114 \\
+26 & 2001/04/22 & 0200 & 2001/04/23 & 1500 & -102 \\
+27 & 2001/08/17 & 1600 & 2001/08/18 & 1600 & -105 \\
+28 & 2001/09/30 & 2300 & 2001/10/02 & 0000 & -148 \\
+29 & 2001/10/21 & 1700 & 2001/10/24 & 1100 & -187 \\
+30 & 2001/10/28 & 0300 & 2001/10/29 & 2200 & -157 \\
+31 & 2002/03/23 & 1400 & 2002/03/25 & 0500 & -100 \\
+32 & 2002/04/17 & 1100 & 2002/04/19 & 0200 & -127 \\
+33 & 2002/04/19 & 0900 & 2002/04/21 & 0600 & -149 \\
+34 & 2002/05/11 & 1000 & 2002/05/12 & 1600 & -110 \\
+35 & 2002/05/23 & 1200 & 2002/05/24 & 2300 & -109 \\
+36 & 2002/08/01 & 2300 & 2002/08/02 & 0900 & -102 \\
+37 & 2002/09/04 & 0100 & 2002/09/05 & 0000 & -109 \\
+38 & 2002/09/07 & 1400 & 2002/09/08 & 2000 & -181 \\
+39 & 2002/10/01 & 0600 & 2002/10/03 & 0800 & -176 \\
+40 & 2002/10/03 & 1000 & 2002/10/04 & 1800 & -146 \\
+41 & 2002/11/20 & 1600 & 2002/11/22 & 0600 & -128 \\
+42 & 2003/05/29 & 2000 & 2003/05/30 & 1000 & -144 \\
+43 & 2003/06/17 & 1900 & 2003/06/19 & 0300 & -141 \\
+44 & 2003/07/11 & 1500 & 2003/07/12 & 1600 & -105 \\
+45 & 2003/08/17 & 1800 & 2003/08/19 & 1100 & -148 \\
+46 & 2003/11/20 & 1200 & 2003/11/22 & 0000 & -422 \\
+47 & 2004/01/22 & 0300 & 2004/01/24 & 0000 & -149 \\
+48 & 2004/02/11 & 1000 & 2004/02/12 & 0000 & -105 \\
+49 & 2004/04/03 & 1400 & 2004/04/04 & 0800 & -112 \\
+50 & 2004/07/22 & 2000 & 2004/07/23 & 2000 & -101 \\
+51 & 2004/07/24 & 2100 & 2004/07/26 & 1700 & -148 \\
+52 & 2004/07/26 & 2200 & 2004/07/30 & 0500 & -197 \\
+53 & 2004/08/30 & 0500 & 2004/08/31 & 2100 & -126 \\
+54 & 2004/11/07 & 2100 & 2004/11/08 & 2100 & -373 \\
+55 & 2004/11/09 & 1100 & 2004/11/11 & 0900 & -289 \\
+56 & 2004/11/11 & 2200 & 2004/11/13 & 1300 & -109 \\
+57 & 2005/01/21 & 1800 & 2005/01/23 & 0500 & -105 \\
+58 & 2005/05/07 & 2000 & 2005/05/09 & 1000 & -127 \\
+59 & 2005/05/29 & 2200 & 2005/05/31 & 0800 & -138 \\
+60 & 2005/06/12 & 1700 & 2005/06/13 & 1900 & -106 \\
+61 & 2005/08/31 & 1200 & 2005/09/01 & 1200 & -131 \\
+62 & 2006/04/13 & 2000 & 2006/04/14 & 2300 & -111 \\
+63 & 2006/12/14 & 2100 & 2006/12/16 & 0300 & -147 \\ \hline
+\end{tabular}%
+\end{table}
+
+\begin{table}[h]
+\centering
+\caption{Evaluation results for models on storm events listed in table \ref{table:teststorms}}
+\label{table:results}
+\begin{tabular}{l c c c}
+\hline
+Model & Mean Absolute Error & Root Mean Square Error & Coefficient of Correlation\\ \hline
+GP-ARX & 7.252 & 11.93 & 0.972\\
+GP-AR & 8.37 & 14.04 & 0.963\\
+Persistence & 9.182 & 14.94 & 0.957\\
+\end{tabular}
+\end{table}
 

### 2017-04-19 chap.tex
Added figures

@@
 We highlight design decisions and practical issues must be addressed in order to build \emph{Gaussian Process} models for probabilistic prediction of a quantity of interest.
-As a case study, we formulate predictive models for one hour ahead prediction of the $Dst$ geomagnetic index.
+As a paedadogical example, we formulate, train and test a family of \emph{Gaussian Process} Auto-Regressive models for one hour ahead prediction of the $Dst$ geomagnetic index.
 
@@
 
+Figures \ref{fig:CompareMae} and \ref{fig:CompareCC} show how the mean absolute error and coefficient of correlation as calculated on the validation set storm events of Table \ref{table:validationstorms}, vary with increasing model order for GP-AR and GP-ARX.
+The results are represented as box and whisker plots, in which a rectangle is drawn to represent the first and third quartiles, with a horizontal line inside to indicate the median value, outlying points are shown as dots while the whiskers indicate the smallest and largest non-outliers.
+In both cases, the predictive performance first improves and then stagnates or worsens with increasing model order.
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{Compare-mae.png}
+\caption{Mean Absolute Error on validation set storms vs model order for GP-AR and GP-ARX.
+\textbf{Key}: Rectangle borders represent the first and third quartiles, with a horizontal line inside to indicate the median value, outlying points are shown as dots and whiskers indicate the smallest and largest non-outliers}
+\label{fig:CompareMae}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{Compare-cc.png}
+\caption{Coefficient of Correlation on validation set storms vs model order for GP-AR and GP-ARX.
+\textbf{Key}: Rectangle borders represent the first and third quartiles, with a horizontal line inside to indicate the median value, outlying points are shown as dots and whiskers indicate the smallest and largest non-outliers}
+\label{fig:CompareCC}
+\end{figure}
+
+Figures \ref{fig:CompareMaeARX} and \ref{fig:CompareCCARX} break down the results for GP-ARX by the model selection routine used.
+Apart from the general trend observed in \ref{fig:CompareMae} and \ref{fig:CompareCC}, we also observe that \emph{grid search} and \emph{coupled simulated annealing} give superior performance as compared to gradient based \emph{maximum likelihood}.
+
+From the validation results, we can chose the model order which yields the best performance, for GP-AR it is $p_t = 6$ while for GP-ARX it is $p_t = 11$.
+Further examination of the validation results shows that in the scheme $p_t = 11$ choosing $p = 7, p_v = 1, p_b = 3$ gives superior results.
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{Compare-mae-arx.png}
+\caption{Mean Absolute Error on validation set storms vs model order for GP-AR and GP-ARX for \emph{CSA}, \emph{GS} and \emph{ML} model selection routines.
+\textbf{Key}: Rectangle borders represent the first and third quartiles, with a horizontal line inside to indicate the median value, outlying points are shown as dots and whiskers indicate the smallest and largest non-outliers}
+\label{fig:CompareMaeARX}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{Compare-cc-arx.png}
+\caption{Coefficient of Correlation on validation set storms vs model order for GP-AR and GP-ARX for \emph{CSA}, \emph{GS} and \emph{ML} model selection routines.
+\textbf{Key}: Rectangle borders represent the first and third quartiles, with a horizontal line inside to indicate the median value, outlying points are shown as dots and whiskers indicate the smallest and largest non-outliers}
+\label{fig:CompareCCARX}
+\end{figure}
+
+After choosing the best performing GP-AR and GP-ARX models, we calculate their performance on the test set of Table \ref{table:teststorms}.
+The results of these model evaluations are summarized in Table \ref{table:results}, the GP-AR and GP-ARX models improve upon the performance of the \emph{persistence model}.
+
+Figures \ref{fig:ComparePred1}, \ref{fig:ComparePred2} and \ref{fig:ComparePred3} show OSA predictions of the GP-ARX model with $\pm \sigma$ error bars for three storm events in the time period between 1998 and 2003.
+The GP-ARX model gives accurate predictions along with plausible error bars around its mean predictions.
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm43.png}
+\caption{OSA Predictions with $\pm \sigma$ error bars for event: 2003/06/17 19:00 to 2003/06/19 03:00}
+\label{fig:ComparePred1}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm8.png}
+\caption{OSA Predictions with $\pm \sigma$ error bars for event: 1998/11/13 00:00 to 1998/11/15 04:00}
+\label{fig:ComparePred2}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm9.png}
+\caption{OSA Predictions with $\pm \sigma$ error bars for event: 1999/01/13 16:00 to 1999/01/14 20:00}
+\label{fig:ComparePred3}
+\end{figure}
+
 \begin{table}[h]
@@
 
+\clearpage
+
 \bibliography{references}

### 2017-04-19 chap.tex
Added subsections to final section

@@
 
-\section{One Hour Ahead $Dst$: Formulation} \label{sec:osa}
+\section{One Hour Ahead $Dst$ Prediction: Formulation} \label{sec:osa}
 
@@
 
-\section{One Hour Ahead $Dst$: Model Design}\label{sec:modeldesign}
+\section{One Hour Ahead $Dst$ Prediction: Model Design}\label{sec:modeldesign}
 
@@
 
-\section{GP-AR and GP-ARX: Workflow Summary}
+\section{GP-AR and GP-ARX: Workflow Summary}\label{sec:workflow}
 
@@
 
+After running the workflow steps as outlined in section \ref{sec:workflow} above, we can understand the performance of the GP-AR and GP-ARX family of models on task of $Dst$ prediction in the following charts.
+
+\subsection*{Model Selection \& Validation Performance}
+
 Figures \ref{fig:CompareMae} and \ref{fig:CompareCC} show how the mean absolute error and coefficient of correlation as calculated on the validation set storm events of Table \ref{table:validationstorms}, vary with increasing model order for GP-AR and GP-ARX.
@@
 
+From the validation results, we can choose the model order which yields the best performance, for GP-AR it is $p_t = 6$ while for GP-ARX it is $p_t = 11$.
+Further examination of the validation results shows that in the scheme $p_t = 11$ choosing $p = 7, p_v = 1, p_b = 3$ gives superior results.
+
+\subsection*{Comparison of Hyper-parameter Selection Algorithms}
 Figures \ref{fig:CompareMaeARX} and \ref{fig:CompareCCARX} break down the results for GP-ARX by the model selection routine used.
 Apart from the general trend observed in \ref{fig:CompareMae} and \ref{fig:CompareCC}, we also observe that \emph{grid search} and \emph{coupled simulated annealing} give superior performance as compared to gradient based \emph{maximum likelihood}.
-
-From the validation results, we can chose the model order which yields the best performance, for GP-AR it is $p_t = 6$ while for GP-ARX it is $p_t = 11$.
-Further examination of the validation results shows that in the scheme $p_t = 11$ choosing $p = 7, p_v = 1, p_b = 3$ gives superior results.
 
@@
 
+\subsection*{Final Evaluation}
+
 After choosing the best performing GP-AR and GP-ARX models, we calculate their performance on the test set of Table \ref{table:teststorms}.
 The results of these model evaluations are summarized in Table \ref{table:results}, the GP-AR and GP-ARX models improve upon the performance of the \emph{persistence model}.
+
+\subsection*{Sample Predictions with Error Bars}
 

### 2017-04-20 chap.tex
Added software section

@@
 In this chapter, we give the reader an indepth view into building of probabilistic forecasting models for geomagnetic time series using the \emph{Gaussian Process} methodology outlined in the previous chapters.
-We highlight design decisions and practical issues must be addressed in order to build \emph{Gaussian Process} models for probabilistic prediction of a quantity of interest.
+We highlight design decisions and practical issues must be addressed in order to use \emph{Gaussian Process} models for probabilistic prediction of a quantity of interest.
 As a paedadogical example, we formulate, train and test a family of \emph{Gaussian Process} Auto-Regressive models for one hour ahead prediction of the $Dst$ geomagnetic index.
@@
 
-\section{Experiments \& Results}
+\section{Practical Issues: Software}
 
-After running the workflow steps as outlined in section \ref{sec:workflow} above, we can understand the performance of the GP-AR and GP-ARX family of models on task of $Dst$ prediction in the following charts.
+Scientific computing, whether it be in the form of simulation, data analysis or modeling is brought to fruition through well thought choices and usage of software packages.
+\emph{Gaussian Processes} have been implemented in a number software distributions, we discuss two of them here.
+
+\begin{enumerate}
+
+\item \textbf{PlasmaML}: An open source project \url{https://gitlab.com/MD-CWI-NL/PlasmaML} maintained at the CWI Amsterdam, it is a software toolbox for Machine Learning applications in Space \& Plasma Physics.
+It is comprised of a number of modules, we use codes from the \textit{omni} module for the experiments and results shown in section \ref{sec:results}.
+It leverages the DynaML machine learning toolkit \url{https://github.com/transcendent-ai-labs/DynaML} which has a well developed library for building GP and Kernel based models and also includes in built support for a number of kernel functions as well as the \emph{Grid Search}, \emph{Maximum Likelihood} and \emph{Coupled Simulated Annealing} model selection routines applied in this chapter.
+\textit{PlasmaML} is written in the \textit{Scala} programming language making it distributable as a \emph{Java Runtime Environment} \textit{jar} file.
+
+\item \textbf{GPML} \citep{GPML}: A \emph{MATLAB} based toolbox for building GP models which has support for a variety of features.
+The software includes a number of kernel functions supported off the shelf, approximate as well as exact inference techniques and support for non-Gaussian likelihood functions.
+\end{enumerate}
+
+\section{Experiments \& Results}\label{sec:results}
+
+After running the workflow steps as outlined in section \ref{sec:workflow} above, we can analyse the performance of the GP-AR and GP-ARX family of models on task of OSA $Dst$ prediction in the following series of charts.
 

### 2017-07-21 revised-chapter.tex
Revisions begun

--- 2017-04-20 chap.tex
@@
 
-In this chapter, we give the reader an indepth view into building of probabilistic forecasting models for geomagnetic time series using the \emph{Gaussian Process} methodology outlined in the previous chapters.
-We highlight design decisions and practical issues must be addressed in order to use \emph{Gaussian Process} models for probabilistic prediction of a quantity of interest.
-As a paedadogical example, we formulate, train and test a family of \emph{Gaussian Process} Auto-Regressive models for one hour ahead prediction of the $Dst$ geomagnetic index.
+In this chapter, we give the reader an in depth view into building of probabilistic forecasting models for geomagnetic time series using the \emph{Gaussian Process} methodology outlined in the previous chapters.
+We highlight design decisions and practical issues that must be addressed in order to use \emph{Gaussian Process} models for probabilistic prediction of a quantity of interest.
+As a pedagogical example, we formulate, train and test a family of \emph{Gaussian Process} Auto-Regressive models for one hour ahead prediction of the $Dst$ geomagnetic index.
 
@@
 It is a region which is impinged upon constantly by the solar wind.
-The ionised plasma ejected by the sun couples with the Earth's magnetic field in a complex manner leading to highly non-linear and chaotic processes which determine the state of the magnetosphere.
+The ionized plasma ejected by the sun couples with the Earth's magnetic field in a complex manner leading to highly non-linear and chaotic processes which determine the state of the magnetosphere.
 It is quite common to reduce these complex dependencies by condensing the state of the Earth's magnetosphere into a set of geomagnetic indices.
@@
     \item $Dst$: A continuous hourly index which gives a measure of the weakening or strengthening of the Earth's equatorial magnetic field due to the weakening or strengthening of the ring currents and the geomagnetic storms \citep{DesslerAndParker}.
-A large negative value of $Dst$ ($ \leq -150 \ nT$) indicates occurence of geomagnetic storm events which are of particular importance in space weather prediction due to their effects on telecommunications infrastructure.
+A large negative value of $Dst$ ($ \leq -150 \ nT$) indicates occurrence of geomagnetic storm events which are of particular importance in space weather prediction due to their effects on telecommunications infrastructure.
 \end{enumerate}
 
-Space weather forecasting systems usually use in-situ measurements of solar wind parameters, taken by satellites, as well as historical data of indices to produce forcasts for various geomagnetic time series.
+Space weather forecasting systems usually use in-situ measurements of solar wind parameters, taken by satellites, as well as historical data of indices to produce forecasts for various geomagnetic time series.
 The principal data source for forecasting of geomagnetic indices is the \emph{OMNI} data set hosted by NASA on the \emph{Space Physics Data Facility} (SPDF).
@@
 Probabilistic forecasting is of particular importance in geophysics applications as the end users of forecasts often require confidence bounds on the said forecasts.
-It is in this context where \emph{Gaussian Processes} become especially attractive due to their inherent probabilistic formulation and tractability of analytical inference.
+It is in this context where \emph{Gaussian Processes} become especially attractive due to their inherent probabilistic formulation and tractability of exact inference.
 
@@
 In the machine learning community, $m(.)$ and $K(.,.)$ are known as the \emph{mean function} and \emph{covariance function} or \emph{kernel function} of the process respectively.
-Giving a closed form expression for $m(.)$ and $K(.,.)$ uniquely specifies a particular \emph{Gaussian Process} and so a GP model is often expressed notationally as shown below.
+Giving a closed form expression for $m(.)$ and $K(.,.)$ uniquely specifies a particular \emph{Gaussian Process} and so a GP model is often expressed in the following notation.
 

### 2017-08-04 revised-chapter.tex
Changes to figures

@@
 
-The Earth's magnetosphere is a region around the planet where the Earth's own magnetic field dominates the magnetic field of outer space.
-It is a region which is impinged upon constantly by the solar wind.
-The ionized plasma ejected by the sun couples with the Earth's magnetic field in a complex manner leading to highly non-linear and chaotic processes which determine the state of the magnetosphere.
+The Earth's magnetosphere is the region above the ionosphere that is defined by the extent of the Earth's magnetic field in space.
+It extends several tens of thousands of kilometers into space, a region which is impinged upon constantly by the solar wind.
+Ionized plasma ejected by the sun couples with the Earth's magnetic field in a complex manner leading to highly non-linear and chaotic processes which determine the state of the magnetosphere.
 It is quite common to reduce these complex dependencies by condensing the state of the Earth's magnetosphere into a set of geomagnetic indices.
@@
 Each geomagnetic index is a proxy for a particular kind of phenomenon.
-Some popular indices are the $K_p$, $Dst$ and the $AE$ index.
+Some common indices are the $Kp$, $Dst$ and the $AE$ index.
 
 \begin{enumerate}
-    \item $K_p$: The Kp-index is a discrete valued global geomagnetic activity index and is based on 3 hour measurements of the K-indices \citep{Bartels}.
-The K-index itself is a three hour long quasi-logarithmic local index of the geomagnetic activity, relative to a calm day curve for the given location.
+    \item $Kp$: The Kp-index is a discrete valued global geomagnetic
+      activity index and is based on 3 hour measurements of the
+      K-indices \citep{Bartels}. The K-index itself is a three hour
+      long quasi-logarithmic local index of the geomagnetic activity,
+      relative to a quiet day curve for the given location.
 
-    \item $AE$: The Auroral Electrojet Index, $AE$, is designed to provide a global, quantitative measure of auroral zone magnetic activity produced by enhanced Ionospheric currents flowing below and within the auroral oval \citep{AEIndex}.
-It is a continuous index which is calculated every hour.
+    \item $AE$: The Auroral Electrojet Index, $AE$, is designed to
+      provide a global, quantitative measure of auroral zone magnetic
+      activity produced by enhanced ionospheric currents flowing below
+      and within the auroral oval \citep{AEIndex}. It is a continuous index which is calculated every hour.
 
-    \item $Dst$: A continuous hourly index which gives a measure of the weakening or strengthening of the Earth's equatorial magnetic field due to the weakening or strengthening of the ring currents and the geomagnetic storms \citep{DesslerAndParker}.
-A large negative value of $Dst$ ($ \leq -150 \ nT$) indicates occurrence of geomagnetic storm events which are of particular importance in space weather prediction due to their effects on telecommunications infrastructure.
+    \item $Dst$: A continuous hourly index which measures the
+      magnitude of the magnetic field induced by near equatorial ring
+      currents \citep{DesslerAndParker}. A large negative value of
+      $Dst$ ($ \leq -100 \ nT$) indicates occurrence of geomagnetic
+      storm events which are of particular importance in space weather
+      prediction due to their adverse effects on telecommunications
+      infrastructure, navigation systems, power grids, satellites etc. 
 \end{enumerate}
 
-Space weather forecasting systems usually use in-situ measurements of solar wind parameters, taken by satellites, as well as historical data of indices to produce forecasts for various geomagnetic time series.
-The principal data source for forecasting of geomagnetic indices is the \emph{OMNI} data set hosted by NASA on the \emph{Space Physics Data Facility} (SPDF).
+Space weather forecasting systems usually use in-situ measurements of
+solar wind parameters, taken by satellites, as well as historical data
+of indices to produce forecasts for various geomagnetic time
+series. Real time data of solar wind parameters are recorded by the \emph{ACE}
+and \emph{DISCOVR} missions. Geomagnetic indices are often recorded in
+ground based measurements. The the \emph{Space Physics Data Facility}
+(SPDF) hosted by NASA is a gateway for accessing data from a variety
+of sources, including those mentioned above.
 
@@
 \subsection{Probabilistic Forecasting}
-%Talk about need for probabilistic forecasts.
+%Talk about need for probabilistic forecasts.ny
 Although much research has been done on prediction of the $Dst$ index, much less has been done on probabilistic forecasting of $Dst$.
@@
 
-In order to uniquely define the distribution of $\mathbf{f}$, it is required to specify $\mathbf{m}$ and $\mathbf{K}$.
-For this probability density to be valid, there are further requirements imposed on $\mathbf{K}$:
+In order to uniquely define the distribution of $\mathbf{f}$, it is
+required to specify the mean $\mathbf{m}$ and the covariance matrix
+$\mathbf{K}$. For this probability density to be valid, there are further requirements imposed on $\mathbf{K}$:
 
@@
 Auto-regressive models can be augmented by including exogenous quantities in the inputs $\mathbf{x}_{t-1}$ at each time step, in order to improve predictive accuracy.
-$Dst$ gives a measure of ring currents, which are modulated by plasma sheet particle injections into the inner magnetosphere during sub-storms.
+$Dst$ gives a measure of ring currents, which are modulated by plasma sheet particle injections into the inner magnetosphere during substorms.
 Studies have shown that the substorm occurrence rate increases with solar wind velocity (high speed streams) \citet{Kissinger2011,Newell2016}.
@@
 
-The quantities $l$ in the RBF, and $b$ and $d$ in the polynomial kernel are known as \emph{hyper-parameters}.
-Hyper-parameters give flexibility to a particular kernel structure, for example $d = 1, 2, 3, \cdots$ in the polynomial kernel represents linear, quadratic, cubic and higher order polynomials respectively.
-The process of assigning values to the \emph{hyper-parameters} is crucial in the model building process and is known as \emph{model selection}.
+The quantities $l$ in the RBF, and $b$ and $d$ in the polynomial kernel are known as \emph{hyperparameters}.
+Hyperparameters give flexibility to a particular kernel structure, for example $d = 1, 2, 3, \cdots$ in the polynomial kernel represents linear, quadratic, cubic and higher order polynomials respectively.
+The process of assigning values to the \emph{hyperparameters} is crucial in the model building process and is known as \emph{model selection}.
 
@@
 
-\subsection{Model Selection: Hyper-parameters}
+\subsection{Model Selection: Hyperparameters}
 
-Given a GPR model with a kernel function $K_\theta$, the problem of model selection consists of finding appropriate values for the kernel hyper-parameters $\theta = \left(\theta_1, \theta_2, \cdots, \theta_i\right)$.
-In order to assign a value to $\theta$, we must define an objective function which represents our confidence that the GP model built from a particular value of $\theta$ is the best performing model.
+Given a GPR model with a kernel function $K_\theta$, the problem of
+model selection consists of finding appropriate values for the kernel
+hyperparameters $\theta = \left(\theta_1, \theta_2, \cdots, \theta_i\right)$.  In order to assign a value to $\theta$, we must
+define an objective function ($Q(\theta)$) which represents our confidence that the GP model built from a particular value of $\theta$ is the best performing model.
 Since GP models are constructed from assumptions about the conditional probability distribution of the data $p(\mathbf{y}|\mathbf{X})$, it is natural to use the negative log-likelihood of the training data as a model selection criterion.
@@
 The objective function $Q(\theta)$ can have multiple local minima, and evaluating the value of $Q(.)$ at any given $\theta$ requires inversion of the matrix $\mathbf{K}_\theta + \sigma^{2} \mathbf{I}$ which has a time complexity $O(N^3)$ as noted above.
-In the interest of saving computational cost, one cannot use exhaustive search through the domain of the hyper-parameters to inform our choice for $\theta$.
+In the interest of saving computational cost, one cannot use exhaustive search through the domain of the hyperparameters to inform our choice for $\theta$.
 
-The model selection problem has historically received less attention due to a combination of difficulty and numerous design decisions like choice of hyper-parameter optimization techniques, which makes model selection as much of an art as science.
-Choosing a suitable heuristic or algorithm for model selection for GP models is quite often a balancing exercise between the computational limitations of the modeler and the need to explore the hyper-parameter space in a manner that yields a satisfactory predictive model.
-Some of the techniques used for model selection in the context of GPR include.
+The model selection problem has historically received less attention
+due to a combination of difficulty and numerous design decisions like
+choice of hyper-parameter optimization techniques, which makes model
+selection as much of an art as science.
+
+Choosing a suitable heuristic or algorithm for model selection for GP
+models is quite often a balancing exercise between the computational
+limitations of the modeler and the need to explore the hyper-parameter
+space in a manner that yields a satisfactory predictive model. Some of
+the techniques used for model selection in the context of GPR include
+the following.
 
@@
 
-Apart from having a practical method for choosing values of kernel hyper-parameters, we must also address the issue of choosing the time delays $p$ and $p_v, p_b$ for the GP-AR and GP-ARX models.
-One might argue that we group these quantities with the kernel hyper-parameters in the model selection procedure, in the context of the GP-AR and GP-ARX models there is one reason why this might not be beneficial.
+Apart from having a practical method for choosing values of kernel hyperparameters, we must also address the issue of choosing the time delays $p$ and $p_v, p_b$ for the GP-AR and GP-ARX models.
+One might argue that we group these quantities with the kernel hyperparameters in the model selection procedure, in the context of the GP-AR and GP-ARX models there is one reason why this might not be beneficial.
 
-The value of kernel hyper-parameters $\theta^*$ which maximise the marginal likelihood $Q(.)$ are conditional on the chosen model orders $p$ and $p, p_v, p_b$ for GP-AR and GP-ARX respectively.
-Since the model orders determine the size of the input space of the predictive models, they influence the values of characteristic length scales and other kernel hyper-parameters which yield a good fit.
+The value of kernel hyperparameters $\theta^*$ which maximise the marginal likelihood $Q(.)$ are conditional on the chosen model orders $p$ and $p, p_v, p_b$ for GP-AR and GP-ARX respectively.
+Since the model orders determine the size of the input space of the predictive models, they influence the values of characteristic length scales and other kernel hyperparameters which yield a good fit.
 
-Not only do the model orders influence the value of kernel hyper-parameters $\theta^*$ which maximise model fit, but increasing values of $p, p_v, p_b$ also make it more difficult for the model selection algorithms to search the hyper-parameter space, due to the curse of dimensionality.
+Not only do the model orders influence the value of kernel hyperparameters $\theta^*$ which maximise model fit, but increasing values of $p, p_v, p_b$ also make it more difficult for the model selection algorithms to search the hyper-parameter space, due to the curse of dimensionality.
 
@@
 We choose a kernel $K_{osa}$ of the form given in equation \ref{eq:usedKernel}.
-In order to find appropriate values of the hyper-parameters of $K_{osa}$, we apply \emph{grid search}, \emph{coupled simulated annealing} and \emph{maximum likelihood} methods.
+In order to find appropriate values of the hyperparameters of $K_{osa}$, we apply \emph{grid search}, \emph{coupled simulated annealing} and \emph{maximum likelihood} methods.
 We fix the parameters $d$ and $\sigma^2$ of $K_{st}$ and model noise to values $0.01$ and $0.2$ respectively, the remaining parameters $w$ and $b$ are kept free to be calculated by model selection.
@@
 $3 \leq p \leq 8$ for GP-AR and $3 \leq p + p_v + p_b \leq 8$ for GP-ARX.
-For each such model perform the model selection routine to select values of the kernel hyper-parameters and evaluate the resultant model on the validation set.
+For each such model perform the model selection routine to select values of the kernel hyperparameters and evaluate the resultant model on the validation set.
 Choose the value of $p, p_v, p_b$ which maximises predictive performance on the validation set.
 
-\item \textbf{Final Evaluation}: After selecting GP-AR and GP-ARX models with particular values of auto-regressive orders and kernel hyper-parameters in the model selection phase, the best performing GP-AR and GP-ARX models are then evaluated on the test data set in order to get an unbiased picture of their predictive performance.
+\item \textbf{Final Evaluation}: After selecting GP-AR and GP-ARX models with particular values of auto-regressive orders and kernel hyperparameters in the model selection phase, the best performing GP-AR and GP-ARX models are then evaluated on the test data set in order to get an unbiased picture of their predictive performance.
 In our case we use the already chosen test set outlined in table \ref{table:teststorms} for final model evaluation.
@@
 \includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm43.png}
-\caption{OSA Predictions with $\pm \sigma$ error bars for event: 2003/06/17 19:00 to 2003/06/19 03:00}
+\caption{OSA Predictions with $\pm \sigma$ error bars for event: 2003/06/17 to 2003/06/19}
 \label{fig:ComparePred1}
@@
 \begin{figure}[h]
-\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm8.png}
-\caption{OSA Predictions with $\pm \sigma$ error bars for event: 1998/11/13 00:00 to 1998/11/15 04:00}
+\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm16.png}
+\caption{OSA Predictions with $\pm \sigma$ error bars for event: 2012/03/08 to 2012/03/10}
 \label{fig:ComparePred2}
@@
 \begin{figure}[h]
-\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm9.png}
-\caption{OSA Predictions with $\pm \sigma$ error bars for event: 1999/01/13 16:00 to 1999/01/14 20:00}
+\includegraphics[width=\textwidth]{PredictionsModel1/PredErrBars_Storm46.png}
+\caption{OSA Predictions with $\pm \sigma$ error bars for event: 2003/11/20 to 2003/11/22}
 \label{fig:ComparePred3}

### 2017-08-07 revised-chapter.tex
Revisions [WIP]

@@
 
-In this chapter, we give the reader an in depth view into building of probabilistic forecasting models for geomagnetic time series using the \emph{Gaussian Process} methodology outlined in the previous chapters.
-We highlight design decisions and practical issues that must be addressed in order to use \emph{Gaussian Process} models for probabilistic prediction of a quantity of interest.
-As a pedagogical example, we formulate, train and test a family of \emph{Gaussian Process} Auto-Regressive models for one hour ahead prediction of the $Dst$ geomagnetic index.
+In this chapter, we give the reader an in depth view into building of
+probabilistic forecasting models for geomagnetic time series using the
+\emph{Gaussian Process} methodology outlined in the previous
+chapters. We highlight design decisions and practical issues that must
+be addressed in order to use \emph{Gaussian Process} models for
+probabilistic prediction of a quantity of interest. As a pedagogical
+example, we formulate, train and test a family of \emph{Gaussian
+  Process} Auto-Regressive models for one hour ahead prediction of the $Dst$ geomagnetic index.
 
@@
 
-\subsection{State of the art}
+\subsection{Models \& Algorithms}
 
@@
 
-\emph{Gaussian Processes} first appeared in machine learning research in \citet{Neal:1996:BLN:525544}, as the limiting case of Bayesian inference performed on neural networks with infinitely many neurons in the hidden layers.
-Although their inception in the machine learning community is recent, their origins can be traced back to the geo-statistics research community where they are known as \emph{Kriging} methods \citep{krige1951statistical}.
-In pure Mathematics, \emph{Gaussian Processes} have been studied extensively and their existence was first proven by Kolmogorov's extension theorem \citep{tao2011introduction}.
-The reader is referred to \citet{Rasmussen:2005:GPM:1162254} for an in depth treatment of Gaussian Processes in machine learning.
+\emph{Gaussian Processes} first appeared in machine learning research
+in \citet{Neal:1996:BLN:525544}, as the limiting case of Bayesian
+inference performed on neural networks with infinitely many neurons in
+the hidden layers. Although their inception in the machine learning
+community is recent, their origins can be traced back to the
+geo-statistics research community where they are known as
+\emph{Kriging} methods \citep{krige1951statistical}. In pure
+Mathematics, \emph{Gaussian Processes} have been studied extensively
+and their existence was first proven by Kolmogorov's extension theorem
+\citep{tao2011introduction}. The reader is referred to
+\citet{Rasmussen:2005:GPM:1162254} for an in depth treatment of
+Gaussian Processes in machine learning. In space weather forecasting
+literature \emph{Gaussian Process Regression} has been used for one
+hour ahead prediction of the Dst index \citep{SWE:SWE20479}.
 
-Without going into too many details, we give a quick recap of the formulation and exact inference in \emph{Gaussian Process Regression} (GPR) models.
+Without going into too many details, we give a quick recap of the
+formulation and exact inference in \emph{Gaussian Process Regression} (GPR) models.
 
@@
 The interested reader is suggested to refer to \citet{Berlinet2004}, \citet{Scholkopf:2001:LKS:559923} and \citet{hofmann2008} for a thorough treatment of kernel functions and the rich theory behind them.
-Some common kernel functions used in machine learning are listed in Table \ref{table:kernel}.
+Some common kernel functions used in machine learning are listed in table \ref{table:kernel}.
 
@@
 
-The quantities $l$ in the RBF, and $b$ and $d$ in the polynomial kernel are known as \emph{hyperparameters}.
+The quantities $l$ in the \emph{Radial Basis Function kernel}, and $b$ and $d$ in the polynomial kernel are known as \emph{hyperparameters}.
 Hyperparameters give flexibility to a particular kernel structure, for example $d = 1, 2, 3, \cdots$ in the polynomial kernel represents linear, quadratic, cubic and higher order polynomials respectively.
@@
 The first training data section consists of ambient fluctuations of $Dst$ while the second contains a geomagnetic storm.
-A set of 24 storm events listed in Table \ref{table:validationstorms} is kept aside as a validation data set for selecting model orders $p, p_v, p_b$ while a list of 63 storm events listed in table \ref{table:teststorms} is used for final model evaluation.
+A set of 24 storm events listed in table \ref{table:validationstorms} is kept aside as a validation data set for selecting model orders $p, p_v, p_b$ while a list of 63 storm events listed in table \ref{table:teststorms} is used for final model evaluation.
 
@@
 
-Figures \ref{fig:CompareMae} and \ref{fig:CompareCC} show how the mean absolute error and coefficient of correlation as calculated on the validation set storm events of Table \ref{table:validationstorms}, vary with increasing model order for GP-AR and GP-ARX.
+Figures \ref{fig:CompareMae} and \ref{fig:CompareCC} show how the mean absolute error and coefficient of correlation as calculated on the validation set storm events of table \ref{table:validationstorms}, vary with increasing model order for GP-AR and GP-ARX.
 The results are represented as box and whisker plots, in which a rectangle is drawn to represent the first and third quartiles, with a horizontal line inside to indicate the median value, outlying points are shown as dots while the whiskers indicate the smallest and largest non-outliers.
@@
 
-After choosing the best performing GP-AR and GP-ARX models, we calculate their performance on the test set of Table \ref{table:teststorms}.
-The results of these model evaluations are summarized in Table \ref{table:results}, the GP-AR and GP-ARX models improve upon the performance of the \emph{persistence model}.
+After choosing the best performing GP-AR and GP-ARX models, we calculate their performance on the test set of table \ref{table:teststorms}.
+The results of these model evaluations are summarized in table \ref{table:results}, the GP-AR and GP-ARX models improve upon the performance of the \emph{persistence model}.
 
@@
 \end{figure}
-
-\begin{table}[h]
-\centering
-\caption{Settings of model selection procedures}
-\begin{tabular}{l c c c}
-\hline
-Procedure & Grid Size & Step & Max Iterations \\
-\hline
-Grid Search & 10 & 0.2 & NA \\
-Coupled Simulated Annealing & 4 & 0.2 & 30 \\
-Maximum likelihood & NA & 0.2 & 150\\
-\end{tabular}
-\label{table:modelselection}
-\end{table}
 
@@
 \centering
+\caption{Settings of model selection procedures}
+\begin{tabular}{l c c c}
+\hline
+Procedure & Grid Size & Step & Max Iterations \\
+\hline
+Grid Search & 10 & 0.2 & NA \\
+Coupled Simulated Annealing & 4 & 0.2 & 30 \\
+Maximum likelihood & NA & 0.2 & 150\\
+\end{tabular}
+\label{table:modelselection}
+\end{table}
+
+\begin{table}[h]
+\centering
 \caption{Evaluation results for models on storm events listed in table \ref{table:teststorms}}

### 2017-08-16 revised-chapter.tex
Revisions

@@
 \begin{document}
+\maketitle
 
@@
 
-%Talk about NARMAX Dst
+% Talk about NARMAX Dst
 Important empirical geomagnetic prediction models include the \emph{Nonlinear Auto-Regessive Moving Average with eXogenous inputs} (NARMAX) methodology (see \citet{doi:10.1080/00207178908559767}, \citet{GRL:GRL13494}, \citet{GRL:GRL20944}, \citet{JGRA:JGRA18657}, \citet{balikhin:narmax}, \citet{JGRA:JGRA20661}, \citet{JGRA:JGRA50192}) and \emph{Artificial Neural Networks} (ANN) based models (\citet{Lund}, \citet{JGRA:JGRA17461}, \citet{SWE:SWE286}, \citet{pallocchia:hal-00318011}) for time series prediction of $Dst$ and $Kp$ indices from interplanetary magnetic field data and solar wind parameters.
@@
 \subsection{Probabilistic Forecasting}
-%Talk about need for probabilistic forecasts.ny
+% Talk about need for probabilistic forecasts.ny
 Although much research has been done on prediction of the $Dst$ index, much less has been done on probabilistic forecasting of $Dst$.
@@
 
-\item \textbf{Model Order Selection}: In order to find appropriate values for the auto-regressive time lags $p, p_v, p_b$ for $Dst$, solar wind speed and IMF $B_z$ respectively, create models upto a certain maximum time lag i.e.
-$3 \leq p \leq 8$ for GP-AR and $3 \leq p + p_v + p_b \leq 8$ for GP-ARX.
-For each such model perform the model selection routine to select values of the kernel hyperparameters and evaluate the resultant model on the validation set.
-Choose the value of $p, p_v, p_b$ which maximises predictive performance on the validation set.
+\item \textbf{Model Order Selection}: In order to find appropriate
+  values for the auto-regressive time lags $p, p_v, p_b$ for $Dst$,
+  solar wind speed and IMF $B_z$ respectively, create models upto a
+  certain maximum time lag i.e. $3 \leq p \leq 12$ for GP-AR and $3
+  \leq p + p_v + p_b \leq 12$ for GP-ARX. For each such model perform
+  the model selection routine to select values of the kernel
+  hyperparameters and evaluate the resultant model on the validation
+  set. Choose the value of $p, p_v, p_b$ which achieves the predictive
+  performance on the validation set (i.e. minimizes the \emph{mean
+    absolute error}).
 
@@
 
-After running the workflow steps as outlined in section \ref{sec:workflow} above, we can analyse the performance of the GP-AR and GP-ARX family of models on task of OSA $Dst$ prediction in the following series of charts.
+After running the workflow steps as outlined in section
+\ref{sec:workflow} above, we are at a position to analyse the
+performance of the GP-AR and GP-ARX family of models on task of OSA
+$Dst$ prediction.
 
 \subsection*{Model Selection \& Validation Performance}
+
+The model validation workflow described in section \ref{sec:workflow}
+produces data containing model validation scores for GP-AR and GP-ARX
+models with different values of autoregressive orders $p_t$. In order
+to better understand the over all trend, we group the performance
+scrores by unique values of $p_t = p + p_v + p_b$ and analyse the
+summary statistics with increasing $p_t$
 
@@
 
-From the validation results, we can choose the model order which yields the best performance, for GP-AR it is $p_t = 6$ while for GP-ARX it is $p_t = 11$.
-Further examination of the validation results shows that in the scheme $p_t = 11$ choosing $p = 7, p_v = 1, p_b = 3$ gives superior results.
+From the charts above, we can choose the model order which yields the
+best \emph{mean absolute error}, for GP-AR it is $p_t = 6$ while for
+GP-ARX it is $p_t = 11$. Further examination of the validation results
+shows that in the scheme $p_t = 11$ choosing $p = 7, p_v = 1, p_b = 3$ gives superior results.
 

### 2017-08-17 revised-chapter.tex
Revisions: Added Conclusion section

@@
 
+Since a GP model expresses prior assumptions about the finite
+dimensional distribution of the quantitiy of interest, the expression
+for $Q(\theta)$ follows directly from the expression of the
+probability density function of a multivariate gaussian.
+
 The model selection problem can now be expressed as the minimization problem shown below.
@@
 
+\section{Conclusion}
+
+In this chapter, we introduced the reader to the \emph{Gaussian
+  Process} prediction methodology in the context of space weather forecasting.
+
+In order to successfully apply a Gaussian Process model for a
+prediction problem, there are a number of design choices which the
+modeler must take. These include choice of model trend, kernel, and
+hyperparameter selection among others, we gave a detailed
+description of each step of the process from the data to a operational
+Gaussian Process forecast model.
+
+To eludicate each step of the modeling workflow, we
+considered as an example, one hour ahead prediction of the Dst
+index.
+
+We demonstrated that in the one hour ahead Dst prediction
+context GP models yield reliable probabilistic predictions with
+relatively small data sizes: the training set contains 243
+instances, while the validation set contains 782 instances.
+
+To understand how well the models perform, their performance is
+compared with the trivial one hour ahead baseline; the
+\emph{persistence model}. The GP-AR and GP-ARX both improved on the
+performance of the \emph{persistence model}.
+
+\emph{Gaussian Processes} present the data driven space weather practitioner with
+a tractable and expressive method for generating probabilistic
+forecasts of geomagnetic time series. Leveraging the strengths of the
+Bayesian approach, they are able to learn robust predictors from data.
+
 \begin{table}[h]

### 2017-08-30 revised-chapter.tex
Formatting changes
 - Moved OMNI to subsection of case study

@@
   example, we formulate, train and test a family of \emph{Gaussian
-    Process} Auto-Regressive models for one hour ahead prediction of the $Dst$ geomagnetic index.
+    Process} Auto-Regressive models for one hour ahead prediction of
+  the $Dst$ geomagnetic index.
 
@@
 
-The Earth's magnetosphere is the region above the ionosphere that is defined by the extent of the Earth's magnetic field in space.
-It extends several tens of thousands of kilometers into space, a region which is impinged upon constantly by the solar wind.
-Ionized plasma ejected by the sun couples with the Earth's magnetic field in a complex manner leading to highly non-linear and chaotic processes which determine the state of the magnetosphere.
-It is quite common to reduce these complex dependencies by condensing the state of the Earth's magnetosphere into a set of geomagnetic indices.
+The Earth's magnetosphere is the region above the ionosphere that is
+defined by the extent of the Earth's magnetic field in space. It
+extends several tens of thousands of kilometers into space, a region
+which is impinged upon constantly by the solar wind. Ionized plasma
+ejected by the sun couples with the Earth's magnetic field in a
+complex manner leading to highly non-linear and chaotic processes
+which determine the state of the magnetosphere. It is quite common to
+reduce these complex dependencies by condensing the state of the
+Earth's magnetosphere into a set of geomagnetic indices.
 
-Geomagnetic indices come in various forms, they may take continuous or discrete values and may be defined with varying time resolutions.
-Their values are often calculated by averaging or combining a number of readings taken by instruments, usually magnetometers, around the Earth.
-Each geomagnetic index is a proxy for a particular kind of phenomenon.
-Some common indices are the $Kp$, $Dst$ and the $AE$ index.
+Geomagnetic indices come in various forms, they may take continuous or
+discrete values and may be defined with varying time
+resolutions. Their values are often calculated by averaging or
+combining a number of readings taken by instruments, usually
+magnetometers, around the Earth. Each geomagnetic index is a proxy for
+a particular kind of phenomenon. Some common indices are the $Kp$,
+$Dst$ and the $AE$ index.
 
@@
   activity produced by enhanced ionospheric currents flowing below
-  and within the auroral oval \citep{AEIndex}. It is a continuous index which is calculated every hour.
+  and within the auroral oval \citep{AEIndex}. It is a continuous
+  index which is calculated every hour.
 
@@
 
-\section{Data Source: OMNI}
-The \emph{OMNI} data set is a collation of in-situ measurements taken by orbiting satellites and ground based measurements of geomagnetic indices.
-The data itself is represented as time averaged versions of the various quantities and is available at two frequencies, by hour and by minute.
-The quantities recorded can be broadly classified into two groups.
-
-\begin{enumerate}
-
-\item Plasma Parameters and Magnetic fields: Measurements of the solar wind particularly the solar wind speed, flow angles, pressure, proton temperature, proton density and field measurements such as strength of electric and magnetic fields as well as their components in various geomagnetic coordinate systems.
-
-\item Indices: $Kp$, $Ae$, $Ap$, $Dst$, sunspot number.
-
-\end{enumerate}
-
-To get a more indepth understanding about the \emph{OMNI} data, interested readers are encouraged to visit the NASA omni web service (\url{https://omniweb.gsfc.nasa.gov}).
-
-In this chapter for the purpose of exposition, we will focus on one hour ahead prediction of the $Dst$ time series.
-From the data we will use time histories of $Dst$, solar wind speed $V$ and the $z$-component of the interplanetary magnetic field $B_z$.
-Our choice is motivated by both physical and practical considerations: solar wind speed and interplanetary magnetic field are important external drivers for particle injection into the magnetosphere and they have large coverage in the \emph{OMNI} data with relatively lesser number of missing records.
+In this chapter for the purpose of exposition, we will focus on one
+hour ahead prediction of the $Dst$ time series. From the data we will
+use time histories of $Dst$, solar wind speed $V$ and the
+$z$-component of the interplanetary magnetic field $B_z$. Our choice
+is motivated by both physical and practical considerations: solar wind
+speed and interplanetary magnetic field are important external drivers
+for particle injection into the magnetosphere and they have large
+coverage in the \emph{OMNI} data with relatively lesser number of
+missing records.
 
@@
 
-A number of modeling techniques have been applied for the prediction of the $Dst$ index.
-One of the earliest forecasting techniques involves calculating the $Dst(t)$ as a solution of an \emph{Ordinary Differential Equation} (ODE) which expressed the rate of change of $Dst(t)$ as a combination of two terms: decay and injection $\frac{d Dst(t)}{dt} = Q(t) - \frac{Dst(t)}{\tau}$, where $Q(t)$ relates to the particle injection from the plasma sheet into the inner magnetosphere.
-This method was presented first by \citet{JGR:JGR10260} and later modified and extended in works such as \citet{Wang:Dst}, \citet{JGRA:JGRA14856}, \citet{Ballatore2014} and others.
+A number of modeling techniques have been applied for the prediction
+of the $Dst$ index. One of the earliest forecasting techniques
+involves calculating the $Dst(t)$ as a solution of an \emph{Ordinary
+  Differential Equation} (ODE) which expressed the rate of change of
+$Dst(t)$ as a combination of two terms: decay and injection $\frac{d
+  Dst(t)}{dt} = Q(t) - \frac{Dst(t)}{\tau}$, where $Q(t)$ relates to
+the particle injection from the plasma sheet into the inner
+magnetosphere. This method was presented first by \citet{JGR:JGR10260}
+and later modified and extended in works such as \citet{Wang:Dst},
+\citet{JGRA:JGRA14856}, \citet{Ballatore2014} and others.
 
 % Talk about NARMAX Dst
-Important empirical geomagnetic prediction models include the \emph{Nonlinear Auto-Regessive Moving Average with eXogenous inputs} (NARMAX) methodology (see \citet{doi:10.1080/00207178908559767}, \citet{GRL:GRL13494}, \citet{GRL:GRL20944}, \citet{JGRA:JGRA18657}, \citet{balikhin:narmax}, \citet{JGRA:JGRA20661}, \citet{JGRA:JGRA50192}) and \emph{Artificial Neural Networks} (ANN) based models (\citet{Lund}, \citet{JGRA:JGRA17461}, \citet{SWE:SWE286}, \citet{pallocchia:hal-00318011}) for time series prediction of $Dst$ and $Kp$ indices from interplanetary magnetic field data and solar wind parameters.
+Important empirical geomagnetic prediction models include the
+\emph{Nonlinear Auto-Regessive Moving Average with eXogenous inputs}
+(NARMAX) methodology (see \citet{doi:10.1080/00207178908559767},
+\citet{GRL:GRL13494}, \citet{GRL:GRL20944}, \citet{JGRA:JGRA18657},
+\citet{balikhin:narmax}, \citet{JGRA:JGRA20661},
+\citet{JGRA:JGRA50192}) and \emph{Artificial Neural Networks} (ANN)
+based models (\citet{Lund}, \citet{JGRA:JGRA17461},
+\citet{SWE:SWE286}, \citet{pallocchia:hal-00318011}) for time series
+prediction of $Dst$ and $Kp$ indices from interplanetary magnetic
+field data and solar wind parameters.
 
@@
 % Talk about need for probabilistic forecasts.ny
-Although much research has been done on prediction of the $Dst$ index, much less has been done on probabilistic forecasting of $Dst$.
-One such work described in \citet{McPherron:2013} involves identification of high speed solar wind streams using the WSA model, using predictions of high speed streams to construct ensembles of $Dst$ trajectories which yield the quartiles of $Dst$ time series.
+Although much research has been done on prediction of the $Dst$ index,
+much less has been done on probabilistic forecasting of $Dst$. One
+such work described in \citet{McPherron:2013} involves identification
+of high speed solar wind streams using the WSA model, using
+predictions of high speed streams to construct ensembles of $Dst$
+trajectories which yield the quartiles of $Dst$ time series.
 
-Probabilistic forecasting is of particular importance in geophysics applications as the end users of forecasts often require confidence bounds on the said forecasts.
-It is in this context where \emph{Gaussian Processes} become especially attractive due to their inherent probabilistic formulation and tractability of exact inference.
+Probabilistic forecasting is of particular importance in geophysics
+applications as the end users of forecasts often require confidence
+bounds on the said forecasts. It is in this context where
+\emph{Gaussian Processes} become especially attractive due to their
+inherent probabilistic formulation and tractability of exact inference.
 
@@
 
-Our aim is to infer an unknown function $f(\mathbf{x})$ from its noise corrupted measurements $(\mathbf{x}_i, y_i)$ where $y_i = f(\mathbf{x}_i) + \epsilon$ and $\epsilon \sim \mathcal{N}(0, \sigma^2)$ is independent and identically distributed Gaussian noise.
+Our aim is to infer an unknown function $f(\mathbf{x})$ from its noise
+corrupted measurements $(\mathbf{x}_i, y_i)$ where $y_i =
+f(\mathbf{x}_i) + \epsilon$ and $\epsilon \sim \mathcal{N}(0,
+\sigma^2)$ is independent and identically distributed Gaussian noise.
 
-A \emph{Gaussian Process} model represents the finite dimensional probability distribution of $f(\mathbf{x}_i)$ by a multivariate gaussian having a particular structure for its mean and covariance as shown in \ref{eq:normal} - \ref{eq:sto}.
+A \emph{Gaussian Process} model represents the finite dimensional
+probability distribution of $f(\mathbf{x}_i)$ by a multivariate
+gaussian having a particular structure for its mean and covariance as
+shown in \ref{eq:normal} - \ref{eq:sto}.
 
@@
 required to specify the mean $\mathbf{m}$ and the covariance matrix
-$\mathbf{K}$. For this probability density to be valid, there are further requirements imposed on $\mathbf{K}$:
+$\mathbf{K}$. For this probability density to be valid, there are
+further requirements imposed on $\mathbf{K}$:
 
@@
 
-In \emph{Gaussian Processes} the individual elements of $\mathbf{x}$ and $\mathbf{K}$ are specified in the form of functions as shown below.
+In \emph{Gaussian Processes} the individual elements of $\mathbf{x}$
+and $\mathbf{K}$ are specified in the form of functions as shown
+below.
 
@@
 
-In the machine learning community, $m(.)$ and $K(.,.)$ are known as the \emph{mean function} and \emph{covariance function} or \emph{kernel function} of the process respectively.
-Giving a closed form expression for $m(.)$ and $K(.,.)$ uniquely specifies a particular \emph{Gaussian Process} and so a GP model is often expressed in the following notation.
+In the machine learning community, $m(.)$ and $K(.,.)$ are known as
+the \emph{mean function} and \emph{covariance function} or
+\emph{kernel function} of the process respectively. Giving a closed
+form expression for $m(.)$ and $K(.,.)$ uniquely specifies a
+particular \emph{Gaussian Process} and so a GP model is often
+expressed in the following notation.
 
@@
 
-In order to generate predictions $f(\mathbf{x}^{*}_i)$ for a set of test points $ {\mathbf{x}^{*}_i : \forall i \in 1, \cdots, M} $.
-Using the multivariate Gaussian distribution in equation (\ref{eq:sto}) we can construct the joint distribution of $f(\mathbf{x})$ over the training and test points.
+In order to generate predictions $f(\mathbf{x}^{*}_i)$ for a set of
+test points $ {\mathbf{x}^{*}_i : \forall i \in 1, \cdots, M} $. Using
+the multivariate Gaussian distribution in equation (\ref{eq:sto}) we
+can construct the joint distribution of $f(\mathbf{x})$ over the
+training and test points.
 
@@
 
-Probabilistic predictions $f_*$ can be generated by constructing the conditional distribution $\mathbf{f_*}|\mathbf{X},\mathbf{y},\mathbf{X_*}$ which is also a multivariate gaussian as shown in equation \ref{eq:posterior}.
+Probabilistic predictions $f_*$ can be generated by constructing the
+conditional distribution
+$\mathbf{f_*}|\mathbf{X},\mathbf{y},\mathbf{X_*}$ which is also a
+multivariate gaussian as shown in equation \ref{eq:posterior}.
 
@@
 
-\section{One Hour Ahead $Dst$ Prediction: Formulation} \label{sec:osa}
+\section{One Hour Ahead $Dst$ Prediction} \label{sec:osa}
 
-Below in equations (\ref{eq:Dst}) - (\ref{eq:DstGP}) we outline a \emph{Gaussian Process} formulation for \emph{OSA} prediction of $Dst$.
-A vector of features $\mathbf{x}_{t-1}$ is used as input to an unknown function $f(\mathbf{x}_{t-1})$.
+\subsection{Data Source: OMNI}
+The \emph{OMNI} data set is a collation of in-situ measurements taken
+by orbiting satellites and ground based measurements of geomagnetic
+indices. The data itself is represented as time averaged versions of
+the various quantities and is available at two frequencies, by hour
+and by minute. The quantities recorded can be broadly classified into
+two groups.
 
-The features $\mathbf{x}_{t-1}$ can be any collection of quantities in the hourly resolution OMNI data set.
-Generally $\mathbf{x}_{t-1}$ are time histories of $Dst$ and other important variables such as plasma pressure $p(t)$, solar wind speed $V(t)$, $z$ component of the interplanetary magnetic field $B_z(t)$.
+\begin{enumerate}
+
+\item Plasma Parameters and Magnetic fields: Measurements of the solar
+  wind particularly the solar wind speed, flow angles, pressure,
+  proton temperature, proton density and field measurements such as
+  strength of electric and magnetic fields as well as their components
+  in various geomagnetic coordinate systems.
+
+\item Indices: $Kp$, $Ae$, $Ap$, $Dst$, sunspot number.
+
+\end{enumerate}
+
+To get a more indepth understanding about the \emph{OMNI} data,
+interested readers are encouraged to visit the NASA omni web service (\url{https://omniweb.gsfc.nasa.gov}).
+
+\subsection{Gaussian Process $Dst$ Model}
+
+Below in equations (\ref{eq:Dst}) - (\ref{eq:DstGP}) we outline a
+\emph{Gaussian Process} formulation for \emph{OSA} prediction of
+$Dst$. A vector of features $\mathbf{x}_{t-1}$ is used as input to an
+unknown function $f(\mathbf{x}_{t-1})$.
+
+The features $\mathbf{x}_{t-1}$ can be any collection of quantities in
+the hourly resolution OMNI data set. Generally $\mathbf{x}_{t-1}$ are
+time histories of $Dst$ and other important variables such as plasma
+pressure $p(t)$, solar wind speed $V(t)$, $z$ component of the
+interplanetary magnetic field $B_z(t)$.
 
@@
 
-We consider two choices for the input features $\mathbf{x}_{t-1}$ leading to two variants of \emph{Gaussian Process} regression for $Dst$ time series prediction.
+We consider two choices for the input features $\mathbf{x}_{t-1}$
+leading to two variants of \emph{Gaussian Process} regression for
+$Dst$ time series prediction.
 
@@
 
-The simplest auto-regressive models for \emph{OSA} prediction of $Dst$ are those that use only the history of $Dst$ to construct input features for model training.
-The input features $\mathbf{x}_{t-1}$ at each time step are the history of $Dst(t)$ until a time lag of $p$ hours.
+The simplest auto-regressive models for \emph{OSA} prediction of $Dst$
+are those that use only the history of $Dst$ to construct input
+features for model training. The input features $\mathbf{x}_{t-1}$ at
+each time step are the history of $Dst(t)$ until a time lag of $p$ hours.
 
@@
 
-Auto-regressive models can be augmented by including exogenous quantities in the inputs $\mathbf{x}_{t-1}$ at each time step, in order to improve predictive accuracy.
-$Dst$ gives a measure of ring currents, which are modulated by plasma sheet particle injections into the inner magnetosphere during substorms.
-Studies have shown that the substorm occurrence rate increases with solar wind velocity (high speed streams) \citet{Kissinger2011,Newell2016}.
-Prolonged southward interplanetary magnetic field (IMF) $z$-component ($B_z$) is needed for sub-storms to occur \citet{McPherron1986}.
-An increase in the solar wind electric field, $V_{sw}B_z$, can increase the dawn-dusk electric field in the magnetotail, which in turn determines the amount of plasma sheet particle that move to the inner magnetosphere \citet{Friedel2001}.
-Therefore, our exogenous parameters consist of solar wind velocity $V_{sw}$ and IMF $B_z$.
+Auto-regressive models can be augmented by including exogenous
+quantities in the inputs $\mathbf{x}_{t-1}$ at each time step, in
+order to improve predictive accuracy. $Dst$ gives a measure of ring
+currents, which are modulated by plasma sheet particle injections into
+the inner magnetosphere during substorms. Studies have shown that the
+substorm occurrence rate increases with solar wind velocity (high
+speed streams) \citet{Kissinger2011,Newell2016}. Prolonged southward
+interplanetary magnetic field (IMF) $z$-component ($B_z$) is needed
+for sub-storms to occur \citet{McPherron1986}. An increase in the
+solar wind electric field, $V_{sw}B_z$, can increase the dawn-dusk
+electric field in the magnetotail, which in turn determines the amount
+of plasma sheet particle that move to the inner magnetosphere
+\citet{Friedel2001}. Therefore, our exogenous parameters consist of
+solar wind velocity $V_{sw}$ and IMF $B_z$.
 
-In this model we choose distinct time lags $p$, $p_{v}$ and $p_{b}$ for $Dst$, $V$ and $B_z$ respectively.
+In this model we choose distinct time lags $p$, $p_{v}$ and $p_{b}$
+for $Dst$, $V$ and $B_z$ respectively.
 
@@
 
-Mean functions in GPR models encode trends in the data, they are the baseline predictions the model falls back to in case the training and test data have little correlation as predicted by the kernel function.
-If there is no prior knowledge about the function to be approximated, \citet{Rasmussen:2005:GPM:1162254} state that it is perfectly reasonable to choose $m(\mathbf{x} = 0)$ as the mean function, as long as the target values are normalized.
-In the case of the $Dst$ time series, it is known that the so called \emph{persistence model} $\hat{D}st(t) = Dst(t-1)$ performs quite well in the context of OSA prediction.
-We therefore choose the \emph{persistence model} as the mean function in our OSA Dst models.
+Mean functions in GPR models encode trends in the data, they are the
+baseline predictions the model falls back to in case the training and
+test data have little correlation as predicted by the kernel
+function. If there is no prior knowledge about the function to be
+approximated, \citet{Rasmussen:2005:GPM:1162254} state that it is
+perfectly reasonable to choose $m(\mathbf{x} = 0)$ as the mean
+function, as long as the target values are normalized. In the case of
+the $Dst$ time series, it is known that the so called
+\emph{persistence model} $\hat{D}st(t) = Dst(t-1)$ performs quite well
+in the context of OSA prediction. We therefore choose the
+\emph{persistence model} as the mean function in our OSA Dst models.
 
@@
 
-For the success of a \emph{Gaussian Process} model an appropriate choice of kernel function is paramount.
-The symmetry and positive semi-definiteness of \emph{Gaussian Process} kernels implies that they represent inner-products between some basis function representation of the data.
-The interested reader is suggested to refer to \citet{Berlinet2004}, \citet{Scholkopf:2001:LKS:559923} and \citet{hofmann2008} for a thorough treatment of kernel functions and the rich theory behind them.
-Some common kernel functions used in machine learning are listed in table \ref{table:kernel}.
+For the success of a \emph{Gaussian Process} model an appropriate
+choice of kernel function is paramount. The symmetry and positive
+semi-definiteness of \emph{Gaussian Process} kernels implies that they
+represent inner-products between some basis function representation of
+the data. The interested reader is suggested to refer to
+\citet{Berlinet2004}, \citet{Scholkopf:2001:LKS:559923} and
+\citet{hofmann2008} for a thorough treatment of kernel functions and
+the rich theory behind them. Some common kernel functions used in
+machine learning are listed in table \ref{table:kernel}.
 
@@
 
-The quantities $l$ in the \emph{Radial Basis Function kernel}, and $b$ and $d$ in the polynomial kernel are known as \emph{hyperparameters}.
-Hyperparameters give flexibility to a particular kernel structure, for example $d = 1, 2, 3, \cdots$ in the polynomial kernel represents linear, quadratic, cubic and higher order polynomials respectively.
-The process of assigning values to the \emph{hyperparameters} is crucial in the model building process and is known as \emph{model selection}.
+The quantities $l$ in the \emph{Radial Basis Function kernel}, and $b$
+and $d$ in the polynomial kernel are known as
+\emph{hyperparameters}. Hyperparameters give flexibility to a
+particular kernel structure, for example $d = 1, 2, 3, \cdots$ in the
+polynomial kernel represents linear, quadratic, cubic and higher order
+polynomials respectively. The process of assigning values to the
+\emph{hyperparameters} is crucial in the model building process and is
+known as \emph{model selection}.
 
-In this text, we construct GPR models with a combination of the \emph{maximum likelihood perceptron} kernel and \emph{student's T} kernel as shown in equations (\ref{eq:usedKernel}).
-The \emph{maximum likelihood perceptron} kernel is the \emph{Gaussian Process} equivalent of a single hidden layer feed-forward neural network model as demonstrated in \citet{Neal:1996:BLN:525544}.
+In this text, we construct GPR models with a combination of the
+\emph{maximum likelihood perceptron} kernel and \emph{student's T}
+kernel as shown in equations (\ref{eq:usedKernel}). The \emph{maximum
+  likelihood perceptron} kernel is the \emph{Gaussian Process}
+equivalent of a single hidden layer feed-forward neural network model
+as demonstrated in \citet{Neal:1996:BLN:525544}.
 
@@
 model selection consists of finding appropriate values for the kernel
-hyperparameters $\theta = \left(\theta_1, \theta_2, \cdots, \theta_i\right)$.  In order to assign a value to $\theta$, we must
-define an objective function ($Q(\theta)$) which represents our confidence that the GP model built from a particular value of $\theta$ is the best performing model.
-Since GP models are constructed from assumptions about the conditional probability distribution of the data $p(\mathbf{y}|\mathbf{X})$, it is natural to use the negative log-likelihood of the training data as a model selection criterion.
+hyperparameters $\theta = \left(\theta_1, \theta_2, \cdots,
+  \theta_i\right)$. In order to assign a value to $\theta$, we must
+define an objective function ($Q(\theta)$) which represents our
+confidence that the GP model built from a particular value of $\theta$
+is the best performing model. Since GP models are constructed from
+assumptions about the conditional probability distribution of the data
+$p(\mathbf{y}|\mathbf{X})$, it is natural to use the negative
+log-likelihood of the training data as a model selection criterion.
 
@@
 
-The model selection problem can now be expressed as the minimization problem shown below.
+The model selection problem can now be expressed as the minimization
+problem shown below.
 
@@
 
-\citet{Rasmussen:2005:GPM:1162254} note that there are ready interpretations to the three terms which add up to give $Q(.)$.
-The first term $\frac{1}{2} \mathbf{y}^\intercal (\mathbf{K}_\theta + \sigma^{2} \mathbf{I})^{-1} \mathbf{y}$ quantifies the fit of the model with respect to the training data, the second term $\frac{1}{2}|\mathbf{K}_\theta + \sigma^{2} \mathbf{I}|$ is a complexity penalty, in non-parametric models such as \emph{Gaussian Processes} the model complexity grows with the number of training data points $N$.
-The third term $\frac{N}{2}log(2\pi)$ is a normalization term.
+\citet{Rasmussen:2005:GPM:1162254} note that there are ready
+interpretations to the three terms which add up to give $Q(.)$. The
+first term $\frac{1}{2} \mathbf{y}^\intercal (\mathbf{K}_\theta +
+\sigma^{2} \mathbf{I})^{-1} \mathbf{y}$ quantifies the fit of the
+model with respect to the training data, the second term
+$\frac{1}{2}|\mathbf{K}_\theta + \sigma^{2} \mathbf{I}|$ is a
+complexity penalty, in non-parametric models such as \emph{Gaussian
+  Processes} the model complexity grows with the number of training
+data points $N$. The third term $\frac{N}{2}log(2\pi)$ is a
+normalization term.
 
-The objective function $Q(\theta)$ can have multiple local minima, and evaluating the value of $Q(.)$ at any given $\theta$ requires inversion of the matrix $\mathbf{K}_\theta + \sigma^{2} \mathbf{I}$ which has a time complexity $O(N^3)$ as noted above.
-In the interest of saving computational cost, one cannot use exhaustive search through the domain of the hyperparameters to inform our choice for $\theta$.
+The objective function $Q(\theta)$ can have multiple local minima, and
+evaluating the value of $Q(.)$ at any given $\theta$ requires
+inversion of the matrix $\mathbf{K}_\theta + \sigma^{2} \mathbf{I}$
+which has a time complexity $O(N^3)$ as noted above. In the interest
+of saving computational cost, one cannot use exhaustive search through
+the domain of the hyperparameters to inform our choice for $\theta$.
 
@@
 
-This is the simplest procedure that may be applied for performing model selection in GPR models.
-This routine constructs a grid of values for $\theta$ as the cartesian product of one dimensional grids for each $\theta_i$, evaluate $Q(.)$ at each such grid point and choose the configuration which yields minimum value of $Q(.)$.
-The advantage of this technique is its simplicity and control over the computational cost of training GP models for practical problems.
-The computational cost of this procedure scales with the total number of points on the grid, this is controlled by the scales and number of grid points for each hyper-parameter $\theta_i$.
+This is the simplest procedure that may be applied for performing
+model selection in GPR models. This routine constructs a grid of
+values for $\theta$ as the cartesian product of one dimensional grids
+for each $\theta_i$, evaluate $Q(.)$ at each such grid point and
+choose the configuration which yields minimum value of $Q(.)$. The
+advantage of this technique is its simplicity and control over the
+computational cost of training GP models for practical problems. The
+computational cost of this procedure scales with the total number of
+points on the grid, this is controlled by the scales and number of
+grid points for each hyper-parameter $\theta_i$.
 
@@
 
-Introduced in \citet{Xavier-De-Souza2010}, \emph{Coupled Simulated Annealing} (CSA) is a family of optimization techniques which is a generalization of \emph{Simulated Annealing} (SA) \citep{Kirkpatrick671}.
-The CSA optimization technique can be understood as a set of parallel SA processes which are coupled by their acceptance probabilities.
+Introduced in \citet{Xavier-De-Souza2010}, \emph{Coupled Simulated
+  Annealing} (CSA) is a family of optimization techniques which is a
+generalization of \emph{Simulated Annealing} (SA)
+\citep{Kirkpatrick671}. The CSA optimization technique can be
+understood as a set of parallel SA processes which are coupled by
+their acceptance probabilities.
 
-From an implementation point of view, CSA follows the same procedure as \emph{grid search}, but after evaluation of the objective $Q(.)$ on the grid, each grid point is iteratively mutated in a random walk fashion.
-This mutation is accepted or rejected according to the new value of $Q(.)$ as well as its value on the other grid points.
-This procedure is iterated until some stop criterion is reached.
+From an implementation point of view, CSA follows the same procedure
+as \emph{grid search}, but after evaluation of the objective $Q(.)$ on
+the grid, each grid point is iteratively mutated in a random walk
+fashion. This mutation is accepted or rejected according to the new
+value of $Q(.)$ as well as its value on the other grid points. This
+procedure is iterated until some stop criterion is reached.
 
-To illustrate how CSA updates are made and accepted consider a population or ensemble of hyper-parameter values ${x_i}$ which can be considered to be initialised in a grid based fashion.
-For each configuration $x_i$ a probing solution $y_i = x_i + \epsilon$ is generated where $\epsilon$ is drawn from some random variable.
-The acceptance probability of the solution is calculated as shown below.
+To illustrate how CSA updates are made and accepted consider a
+population or ensemble of hyper-parameter values ${x_i}$ which can be
+considered to be initialised in a grid based fashion. For each
+configuration $x_i$ a probing solution $y_i = x_i + \epsilon$ is
+generated where $\epsilon$ is drawn from some random variable. The
+acceptance probability of the solution is calculated as shown below.
 
@@
 
-The quantity $\gamma$ defines the \emph{coupling} between the states ${x_i}$ of the ensemble, when the number of configurations in the ensemble $x_i$ is one this rule reduces to the classical \emph{Metropolis-Hastings} acceptance rule of the SA algorithm.
-Up to 4 variants of the CSA algorithm have been defined with different acceptance rules and temperature annealing schedules, the reader is advised to refer to \citet{Xavier-De-Souza2010} for an in-depth treatment of the CSA optimization class.
+The quantity $\gamma$ defines the \emph{coupling} between the states
+${x_i}$ of the ensemble, when the number of configurations in the
+ensemble $x_i$ is one this rule reduces to the classical
+\emph{Metropolis-Hastings} acceptance rule of the SA algorithm. Up to
+4 variants of the CSA algorithm have been defined with different
+acceptance rules and temperature annealing schedules, the reader is
+advised to refer to \citet{Xavier-De-Souza2010} for an in-depth
+treatment of the CSA optimization class.
 
@@
 
-This technique, as described in \citet{Rasmussen:2005:GPM:1162254}, is a form of \emph{gradient descent}.
-It involves starting with an initial guess for $\theta$ and iteratively improving it by calculating the gradient of $Q(.)$ with respect to $\theta$.
-The gradient of $Q(.)$ with respect to each hyper-parameter $\theta_i$ can be calculated analytically as shown below.
+This technique, as described in \citet{Rasmussen:2005:GPM:1162254}, is
+a form of \emph{gradient descent}. It involves starting with an
+initial guess for $\theta$ and iteratively improving it by calculating
+the gradient of $Q(.)$ with respect to $\theta$. The gradient of
+$Q(.)$ with respect to each hyper-parameter $\theta_i$ can be
+calculated analytically as shown below.
 
@@
 
-Although gradient based \emph{Maximum Likelihood} looks very similar to the \emph{Gradient Descent} used to learn parametric regression models like \emph{Feed forward Neural Networks}, in the GPR model selection context it introduces an extra computational cost of calculating the gradient of $Q(\theta)$ with respect to each $\theta_i$ in every iteration.
-Applying this method can sometimes lead to overfitting of the GPR model to the training data \citep{Rasmussen:2005:GPM:1162254}.
+Although gradient based \emph{Maximum Likelihood} looks very similar
+to the \emph{Gradient Descent} used to learn parametric regression
+models like \emph{Feed forward Neural Networks}, in the GPR model
+selection context it introduces an extra computational cost of
+calculating the gradient of $Q(\theta)$ with respect to each
+$\theta_i$ in every iteration. Applying this method can sometimes lead
+to overfitting of the GPR model to the training data \citep{Rasmussen:2005:GPM:1162254}.
 
@@
 
-Apart from having a practical method for choosing values of kernel hyperparameters, we must also address the issue of choosing the time delays $p$ and $p_v, p_b$ for the GP-AR and GP-ARX models.
-One might argue that we group these quantities with the kernel hyperparameters in the model selection procedure, in the context of the GP-AR and GP-ARX models there is one reason why this might not be beneficial.
+Apart from having a practical method for choosing values of kernel
+hyperparameters, we must also address the issue of choosing the time
+delays $p$ and $p_v, p_b$ for the GP-AR and GP-ARX models. One might
+argue that we group these quantities with the kernel hyperparameters
+in the model selection procedure, in the context of the GP-AR and
+GP-ARX models there is one reason why this might not be beneficial.
 
-The value of kernel hyperparameters $\theta^*$ which maximise the marginal likelihood $Q(.)$ are conditional on the chosen model orders $p$ and $p, p_v, p_b$ for GP-AR and GP-ARX respectively.
-Since the model orders determine the size of the input space of the predictive models, they influence the values of characteristic length scales and other kernel hyperparameters which yield a good fit.
+The value of kernel hyperparameters $\theta^*$ which maximise the
+marginal likelihood $Q(.)$ are conditional on the chosen model orders
+$p$ and $p, p_v, p_b$ for GP-AR and GP-ARX respectively. Since the
+model orders determine the size of the input space of the predictive
+models, they influence the values of characteristic length scales and
+other kernel hyperparameters which yield a good fit.
 
-Not only do the model orders influence the value of kernel hyperparameters $\theta^*$ which maximise model fit, but increasing values of $p, p_v, p_b$ also make it more difficult for the model selection algorithms to search the hyper-parameter space, due to the curse of dimensionality.
+Not only do the model orders influence the value of kernel
+hyperparameters $\theta^*$ which maximise model fit, but increasing
+values of $p, p_v, p_b$ also make it more difficult for the model
+selection algorithms to search the hyper-parameter space, due to the
+curse of dimensionality.
 
-An empirical way to choose the values of $p, p_v, p_b$ for the GP-AR and GP-ARX models is by choosing an independent validation data set which is used to evaluate the GP models obtained from the model selection procedures of the previous section.
-It is then possible for the modeler to choose these model orders in way to balance performance and dimensionality of input features.
+An empirical way to choose the values of $p, p_v, p_b$ for the GP-AR
+and GP-ARX models is by choosing an independent validation data set
+which is used to evaluate the GP models obtained from the model
+selection procedures of the previous section. It is then possible for
+the modeler to choose these model orders in way to balance performance
+and dimensionality of input features.
 
@@
 
-Section \ref{sec:modeldesign} brings together all the practical issues with respect to building of GP-AR and GP-ARX models for predicting $Dst$, they can be condensed to form the following steps.
+Section \ref{sec:modeldesign} brings together all the practical issues
+with respect to building of GP-AR and GP-ARX models for predicting
+$Dst$, they can be condensed to form the following steps.
 
@@
 
-\item \textbf{Data Preprocessing}: Three seperate data sets for model training and hyper-parameter selection, model order selection and model evaluation must be created.
-Care must be taken that the size of the training sets does not delay the processes of kernel hyper-parameter selection.
-The training, validation and test data sets created from this procedure must all contain non-overlapping geomagnetic storm events.
+\item \textbf{Data Preprocessing}: Three seperate data sets for model
+  training and hyper-parameter selection, model order selection and
+  model evaluation must be created. Care must be taken that the size
+  of the training sets does not delay the processes of kernel
+  hyper-parameter selection. The training, validation and test data
+  sets created from this procedure must all contain non-overlapping
+  geomagnetic storm events.
 
-  We selected OMNI data sections 00:00 January 3 2010 to 23:00 January 23 2010 and 20:00 August 5 2011 to 22:00 August 6 2011 for training the GP-AR and GP-ARX models.
-The first training data section consists of ambient fluctuations of $Dst$ while the second contains a geomagnetic storm.
-A set of 24 storm events listed in table \ref{table:validationstorms} is kept aside as a validation data set for selecting model orders $p, p_v, p_b$ while a list of 63 storm events listed in table \ref{table:teststorms} is used for final model evaluation.
+  We selected OMNI data sections 00:00 January 3 2010 to 23:00 January
+  23 2010 and 20:00 August 5 2011 to 22:00 August 6 2011 for training
+  the GP-AR and GP-ARX models. The first training data section
+  consists of ambient fluctuations of $Dst$ while the second contains
+  a geomagnetic storm. A set of 24 storm events listed in table
+  \ref{table:validationstorms} is kept aside as a validation data set
+  for selecting model orders $p, p_v, p_b$ while a list of 63 storm
+  events listed in table \ref{table:teststorms} is used for final
+  model evaluation.
 
@@
 
-  We choose a kernel $K_{osa}$ of the form given in equation \ref{eq:usedKernel}.
-In order to find appropriate values of the hyperparameters of $K_{osa}$, we apply \emph{grid search}, \emph{coupled simulated annealing} and \emph{maximum likelihood} methods.
-We fix the parameters $d$ and $\sigma^2$ of $K_{st}$ and model noise to values $0.01$ and $0.2$ respectively, the remaining parameters $w$ and $b$ are kept free to be calculated by model selection.
-Table \ref{table:modelselection} summarizes the settings used to run each model selection procedure.
+  We choose a kernel $K_{osa}$ of the form given in equation
+  \ref{eq:usedKernel}. In order to find appropriate values of the
+  hyperparameters of $K_{osa}$, we apply \emph{grid search},
+  \emph{coupled simulated annealing} and \emph{maximum likelihood}
+  methods. We fix the parameters $d$ and $\sigma^2$ of $K_{st}$ and
+  model noise to values $0.01$ and $0.2$ respectively, the remaining
+  parameters $w$ and $b$ are kept free to be calculated by model
+  selection. Table \ref{table:modelselection} summarizes the settings
+  used to run each model selection procedure.
 
-\item \textbf{Choice of Performance Metrics}: To compare the performance of GP and GP-ARX models of varying auto-regressive orders, their predictive performance during the validation (and later final evaluation) phase must be calculated using some clearly defined metrics.
-We use the following as performance metrics for our modeling experiments.
+\item \textbf{Choice of Performance Metrics}: To compare the
+  performance of GP and GP-ARX models of varying auto-regressive
+  orders, their predictive performance during the validation (and
+  later final evaluation) phase must be calculated using some clearly
+  defined metrics. We use the following as performance metrics for our
+  modeling experiments.
 
@@
 
-\item \textbf{Final Evaluation}: After selecting GP-AR and GP-ARX models with particular values of auto-regressive orders and kernel hyperparameters in the model selection phase, the best performing GP-AR and GP-ARX models are then evaluated on the test data set in order to get an unbiased picture of their predictive performance.
-In our case we use the already chosen test set outlined in table \ref{table:teststorms} for final model evaluation.
-The GP based OSA prediction models are compared against the naive \emph{Persistence} model which serves as a performance baseline for the OSA prediction of $Dst$.
+\item \textbf{Final Evaluation}: After selecting GP-AR and GP-ARX
+  models with particular values of auto-regressive orders and kernel
+  hyperparameters in the model selection phase, the best performing
+  GP-AR and GP-ARX models are then evaluated on the test data set in
+  order to get an unbiased picture of their predictive performance. In
+  our case we use the already chosen test set outlined in table
+  \ref{table:teststorms} for final model evaluation. The GP based OSA
+  prediction models are compared against the naive \emph{Persistence}
+  model which serves as a performance baseline for the OSA prediction
+  of $Dst$.
 
@@
 
-Scientific computing, whether it be in the form of simulation, data analysis or modeling is brought to fruition through well thought choices and usage of software packages.
-\emph{Gaussian Processes} have been implemented in a number software distributions, we discuss two of them here.
+Scientific computing, whether it be in the form of simulation, data
+analysis or modeling is brought to fruition through well thought
+choices and usage of software packages. \emph{Gaussian Processes} have
+been implemented in a number software distributions, we discuss two of
+them here.
 
@@
 
-\item \textbf{PlasmaML}: An open source project \url{https://gitlab.com/MD-CWI-NL/PlasmaML} maintained at the CWI Amsterdam, it is a software toolbox for Machine Learning applications in Space \& Plasma Physics.
-It is comprised of a number of modules, we use codes from the \textit{omni} module for the experiments and results shown in section \ref{sec:results}.
-It leverages the DynaML machine learning toolkit \url{https://github.com/transcendent-ai-labs/DynaML} which has a well developed library for building GP and Kernel based models and also includes in built support for a number of kernel functions as well as the \emph{Grid Search}, \emph{Maximum Likelihood} and \emph{Coupled Simulated Annealing} model selection routines applied in this chapter.
-\textit{PlasmaML} is written in the \textit{Scala} programming language making it distributable as a \emph{Java Runtime Environment} \textit{jar} file.
+\item \textbf{PlasmaML}: An open source project
+  \url{https://gitlab.com/MD-CWI-NL/PlasmaML} maintained at the CWI
+  Amsterdam, it is a software toolbox for Machine Learning
+  applications in Space \& Plasma Physics. It is comprised of a number
+  of modules, we use codes from the \textit{omni} module for the
+  experiments and results shown in section \ref{sec:results}. It
+  leverages the DynaML machine learning toolkit
+  \url{https://github.com/transcendent-ai-labs/DynaML} which has a
+  well developed library for building GP and Kernel based models and
+  also includes in built support for a number of kernel functions as
+  well as the \emph{Grid Search}, \emph{Maximum Likelihood} and
+  \emph{Coupled Simulated Annealing} model selection routines applied
+  in this chapter. \textit{PlasmaML} is written in the \textit{Scala}
+  programming language making it distributable as a \emph{Java Runtime
+    Environment} \textit{jar} file.
 
-\item \textbf{GPML} \citep{GPML}: A \emph{MATLAB} based toolbox for building GP models which has support for a variety of features.
-The software includes a number of kernel functions supported off the shelf, approximate as well as exact inference techniques and support for non-Gaussian likelihood functions.
+\item \textbf{GPML} \citep{GPML}: A \emph{MATLAB} based toolbox for
+  building GP models which has support for a variety of features. The
+  software includes a number of kernel functions supported off the
+  shelf, approximate as well as exact inference techniques and support
+  for non-Gaussian likelihood functions. 
 \end{enumerate}
@@
 
-Figures \ref{fig:CompareMae} and \ref{fig:CompareCC} show how the mean absolute error and coefficient of correlation as calculated on the validation set storm events of table \ref{table:validationstorms}, vary with increasing model order for GP-AR and GP-ARX.
-The results are represented as box and whisker plots, in which a rectangle is drawn to represent the first and third quartiles, with a horizontal line inside to indicate the median value, outlying points are shown as dots while the whiskers indicate the smallest and largest non-outliers.
-In both cases, the predictive performance first improves and then stagnates or worsens with increasing model order.
+Figures \ref{fig:CompareMae} and \ref{fig:CompareCC} show how the mean
+absolute error and coefficient of correlation as calculated on the
+validation set storm events of table \ref{table:validationstorms},
+vary with increasing model order for GP-AR and GP-ARX. The results are
+represented as box and whisker plots, in which a rectangle is drawn to
+represent the first and third quartiles, with a horizontal line inside
+to indicate the median value, outlying points are shown as dots while
+the whiskers indicate the smallest and largest non-outliers. In both
+cases, the predictive performance first improves and then stagnates or
+worsens with increasing model order.
 
@@
 \subsection*{Comparison of Hyper-parameter Selection Algorithms}
-Figures \ref{fig:CompareMaeARX} and \ref{fig:CompareCCARX} break down the results for GP-ARX by the model selection routine used.
-Apart from the general trend observed in \ref{fig:CompareMae} and \ref{fig:CompareCC}, we also observe that \emph{grid search} and \emph{coupled simulated annealing} give superior performance as compared to gradient based \emph{maximum likelihood}.
+Figures \ref{fig:CompareMaeARX} and \ref{fig:CompareCCARX} break down
+the results for GP-ARX by the model selection routine used. Apart from
+the general trend observed in \ref{fig:CompareMae} and
+\ref{fig:CompareCC}, we also observe that \emph{grid search} and
+\emph{coupled simulated annealing} give superior performance as
+compared to gradient based \emph{maximum likelihood}.
 
@@
 
-After choosing the best performing GP-AR and GP-ARX models, we calculate their performance on the test set of table \ref{table:teststorms}.
-The results of these model evaluations are summarized in table \ref{table:results}, the GP-AR and GP-ARX models improve upon the performance of the \emph{persistence model}.
+After choosing the best performing GP-AR and GP-ARX models, we
+calculate their performance on the test set of table
+\ref{table:teststorms}. The results of these model evaluations are
+summarized in table \ref{table:results}, the GP-AR and GP-ARX models
+improve upon the performance of the \emph{persistence model}.
 
@@
 
-Figures \ref{fig:ComparePred1}, \ref{fig:ComparePred2} and \ref{fig:ComparePred3} show OSA predictions of the GP-ARX model with $\pm \sigma$ error bars for three storm events in the time period between 1998 and 2003.
-The GP-ARX model gives accurate predictions along with plausible error bars around its mean predictions.
+Figures \ref{fig:ComparePred1}, \ref{fig:ComparePred2} and
+\ref{fig:ComparePred3} show OSA predictions of the GP-ARX model with
+$\pm \sigma$ error bars for three storm events in the time period
+between 1998 and 2003. The GP-ARX model gives accurate predictions
+along with plausible error bars around its mean predictions.
 

### 2017-08-30 response-reviewer1.tex
Revision: [WIP]

\documentclass{article}
\begin{document}

\title{Response Reviewer 1}
\maketitle

The author would like to thank the reviewer for the various comments
and suggestions which greatly helped in improving the quality of the manuscript.

In the proceeding section they will find the revisions/responses for
each concern outlined. The tracked changes are highlighted in the file \emph{differences.pdf}.

\section*{Draft Comments \& Revisions}

\begin{enumerate}

\item{

\fbox{%
    \parbox{\textwidth}{%
\textbf{Comment}: The abstract assumes that the readers are already introduced to the GP methodology from previous chapters.}
}

\textbf{Response}: I concur with the assertion, given the scope and
organisation of the text it is not an unreasonable assumption. The
chapter nevertheless strives to give a bare minimum theoretical
treatment of Gaussian Process modelelling.

}

\item{
  \fbox{%
    \parbox{\textwidth}{%
\textbf{Comment}: Replace $K_p$ with Kp}
}

\textbf{Response}: Appropriate changes made in the manuscript.

}

\item{

\fbox{%
\parbox{\textwidth}{%
\textbf{Comment}: Section 1, last paragraph. The SPDF is a repository
for a wide variety of data sources from spacecrafts and ground-based
instruments. It is not intended to forecast geomagnetic indices. It is
intended for research and validation purposes only.
}
}

\textbf{Response}: Appropriate changes made in section 1, page 2.

}

\item{

\fbox{%
\parbox{\textwidth}{%

\textbf{Comment}: Section 3. Can the author point to exactly why the
Dst is chosen for this work as opposed to some other ground-based
index? Perhaps also identify other tools and techniques for the
prediction of other indices similar to the Dst. Also “state of the
art” could be better said “other forecast alogrithms/models”. Some of
the references are obsolete while many of them are unverified and have
not transitioned to forecast models. The notable ones are the Wing et
al. 2005 and Bala et al. 2009 which are operational in near-real
time. Wing et al. 2005 was picked by NOAA SWPC for operational purpose.

}
}

\textbf{Response}: Since scope of the chapter is to illustrate how to
apply GP methodology for space weather forecasting, the particular
choice of geomagnetic index and its justification is tangential to the
primary aim. That being said, it is evident from the introduction that
the Dst index being a measure of geomagnetic disturbance strength is
an important quantity and thus an easy choice for a case study.

The changes with respect to forecast algorithms has been made on page
3, section 2.1.
}

\item{

\fbox{%
\parbox{\textwidth}{%
\textbf{Comment}: Section 3.2: more examples of probabilistic
forecasting models in the space weather context would be relevant here.

}
}

\textbf{Response}: Changes made on page 3, section 2.2.

}

\item{

\fbox{%
\parbox{\textwidth}{%
\textbf{Comment}: Section 4... Are there references in the space
weather literature with applications involving this technique?
}
}

\textbf{Response}: A reference using the current technique has been
added on page 4, section 4.

}

\item{

\fbox{%
\parbox{\textwidth}{

\textbf{Comment}: Section 4.1: Given this will cover a broader
audience, a sentence or two introducing the GP to the ML context would
be relevant. Perhaps saying it as “this is a kernel-based, supervised
learning approach that is good for prediction tasks” or such could
make a nice segue into the section.

}
}

\textbf{Response}: Gaussian Processes have a rich theory intersecting
areas of probability theory, functional analysis, linear
operators and positive type functions. Because the multiple theoretical
perspectives that can be used to introduce them, there is a chance of
creating confusion for new practitioners especially those with less 
mathematical experience. Thus it was deemed non-essential to mention
terms such as "kernel-based, supervised learning approaches". The
chapter nevertheless gives a basic introduction to kernel functions
and their characteristics.

}

\item{
    \fbox{
      \parbox{\textwidth}{
        \textbf{Comment}: Section 5: define OSA.
$X_{t-1}$ having composed of the predicted index in the formulation can be
precarious and can have some unintended consequences as a result. Dst
is a slow response index to the changing IMF conditions. Applying time
histories of the Dst index along with the prevailing IMF parameters
will be an interesting prediction model at the outset- two causally
coupled quantities with a large time lag (>1 hr) as inputs. It would
be interesting to see a mere persistence model like the GPR comparing
with this. It will also be nice to see a model that only takes the SW
as inputs to predict the Dst index.

      }
    }
    \textbf{Response}: The persistence model has been compared with
    the GP based models in table 5 on page 23. Due to strong
    persistence behavior, the author believes that a model based
    solely on solar wind inputs would not give attrative
    performance. For that reason and for sake of brevity it has been
    omitted from the chapter.

}

\item{
    \fbox{
      \parbox{\textwidth}{
        \textbf{Comment}: Section 5.2: The author has chosen Vsw and
        Bz to augment their original AR model. In predicting the Dst
        index it is useful to bring in the dynamic pressure term or at
        a minimum the density term from the SW parameters. Did the
        author try to model that? It will be good to see the density
        term added in their or a mention of why it was omitted. A
        clarification is needed here.

      }
    }
    \textbf{Response}: Modifications made on page 6, section 4.4.

}

\item{
    \fbox{
      \parbox{\textwidth}{
        \textbf{Comment}: Section 7.4, It is not clear how the GPARX
        model number is chosen from the conditions shown here: 3 <= p
        + pv + pb <= 8. However the validation shows that this is not
        preserved. Can the author explain this?

      }
    }
    \textbf{Response}: Modifications made on page 12, section 6.

}

\item{
    \fbox{
      \parbox{\textwidth}{
        \textbf{Comment}: Figures 5 and 6 tend to indicate the problem
        of persistence alluded to in my comment 8 above. It looks
        (hard to tell from a b/w figure though) as if the prediction
        is following the actual Dst. If this is indeed the case (again
        please clarify this with a better plot or with color), the
        model clearly misses the onset which is critical to a good
        forecast model.

      }
    }
    \textbf{Response}: Figures 5, 6 and 7 have been modified, also
    effort was made to include a strong geomagnetic event in them.

}

\item{
    \fbox{
      \parbox{\textwidth}{
        \textbf{Comment}: How does the results here compare to other
        models of similar technique? Can the author also talk about
        noise free observations if there is relevance here?

      }
    }
    \textbf{Response}: Since the noise component in the model has
    small variance, it is very close to the noise free paradigm. Since
    is difficult to obtain codes for various space weather models and
    since it is not the aim of the chapter to compare forecast models,
    comparisons with other methods have been avoided. It is also worth
    noting that a very large number of Dst forecast models are
    deterministic forecasts and not probabilistic.

}

\item{
    \fbox{
      \parbox{\textwidth}{
        \textbf{Comment}: Even though it may not be relevant here it
        might be good to talk about the operational scalability of
        this model to a real-time framework and any other key
        considerations from a forecaster perspective.

      }
    }
    \textbf{Response}: Although operational scalability of the
    technique is not of relevance and not explicitly discussed,
    on page 8 the reader is informed on the time complexity of GP
    based inference.

}

\item{
    \fbox{
      \parbox{\textwidth}{
        \textbf{Comment}: The title “many hours ahead prediction..”,
        at least in my view, is slightly misleading. Generally this
        refers to long-term predictions. But I find this as an
        artifact of the GP process which only comes with familiarity
        of the process itself.

      }
    }
    \textbf{Response}: The title "many hours ahead prediction .." was
    a working title and has been modified when submitting the first
    draft of the chapter. Due to the online editorial system not being
    able to reflect that modification there was a
    miscommunication. The revised draft has an accurate chapter title.

}

\end{enumerate}

\end{document}

### 2017-08-30 revised-chapter.tex
Revision: [WIP]

@@
 of indices to produce forecasts for various geomagnetic time
-series. Real time data of solar wind parameters are recorded by the \emph{ACE}
-and \emph{DISCOVR} missions. Geomagnetic indices are often recorded in
+series. Geomagnetic indices are often recorded in
 ground based measurements. The the \emph{Space Physics Data Facility}
 (SPDF) hosted by NASA is a gateway for accessing data from a variety
-of sources, including those mentioned above.
+of sources, including those mentioned above, it is meant for use in
+research and validation. In order to deploy space weather prediction
+systems, real time data of solar wind parameters can be obtained from
+the \emph{ACE} and \emph{DISCOVR} missions.
 
@@
 predictions of high speed streams to construct ensembles of $Dst$
-trajectories which yield the quartiles of $Dst$ time series.
+trajectories which yield the quartiles of $Dst$ time
+series. \citet{mcpherron2004probabilistic} use a similar methodology
+for probabilistic forecasting of the 3-hour ahead $Ap$ index.
 
@@
 of plasma sheet particle that move to the inner magnetosphere
-\citet{Friedel2001}. Therefore, our exogenous parameters consist of
-solar wind velocity $V_{sw}$ and IMF $B_z$.
+\citet{Friedel2001}.
 
-In this model we choose distinct time lags $p$, $p_{v}$ and $p_{b}$
-for $Dst$, $V$ and $B_z$ respectively.
+Although there are also other solar wind inputs such as dynamic
+pressure $p$ and proton density $\rho$, it is known that the dynamic
+pressure  is a function of the solar wind velocity and the particle
+densities. Further there are gaps in the OMNI data for these
+quantities making them less attractive as compared to the solar wind
+velocity and the IMF data.
+
+Therefore, our exogenous parameters consist of solar wind velocity
+$V_{sw}$ and IMF $B_z$. We choose distinct time lags $p$, $p_{v}$ and
+$p_{b}$ for $Dst$, $V$ and $B_z$ respectively.
 
@@
     absolute error}).
+
+  The reader should remember that range of time
+  lags to search ($3 \leq p \leq 12$) is a balance between
+  computational effort and predictive performance. This choice is
+  application specific and can be better informed by exploratory
+  analysis of the geomagnetic time series.
 

### 2017-08-31 response-reviewer2.tex
Revised draft

\documentclass{article}
\begin{document}

\title{Response Reviewer 2}
\maketitle

The author would like to thank the reviewer for the various comments
and suggestions which greatly helped in improving the quality of the manuscript.

In the proceeding section they will find the revisions/responses for
each concern outlined. The tracked changes are highlighted in the file \emph{differences.pdf}.

\section*{Draft Comments \& Revisions}

\begin{enumerate}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: Within the abstract the word ‘that’ is missing
      from “We highlight design decisions and practical issues that
      must be addressed...”

    }
  }

\textbf{Response}: Modifications done.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: When describing the magnetosphere, saying that
      Earth’s field “dominates the magnetic field of outer space” is
      vague. “Outer space” could be defined as outside the
      heliosphere. Within the heliosphere the sun’s magnetic field is
      dominant except for close to planets with their own planetary magnetospheres.
    }
  }

\textbf{Response}: Changes made in section 1 on page 1.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: I would suggest replacing “some popular
      indices are” with “some common indices are.”
    }
  }

\textbf{Response}: Changes done.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: Typically we would say “quiet day curve”
      instead of “calm day curve”
    }
  }

\textbf{Response}: Changes made on page 1.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: Ionosphere does not need to be capitalized in
      the description of the AE index
    }
  }

\textbf{Response}: Changes made on page 2.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: Reword the Dst section as the first sentence
      is confusion and difficult to read. Perhaps omitting “weakening
      or strengthening” and just discussing how Dst measures the
      strength of the ring current.
    }
  }

\textbf{Response}: Changes made on page 2.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: A geomagnetic storm with a Dst of ≤ 150 nT
      would be quiet large. Typically smaller values are used to
      indicate the occurrence of the geomagnetic storm (value of ≤ 100
      nT or so). It might be good to include a reference here.
    }
  }

\textbf{Response}: Changes made on page 2.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: The space weather effects of geomagnetic
      storms are much greater than just “their effects on
      telecommunications infrastructure”. There are lots of other
      effects that should be mentioned as well: the effects of the
      power grid, satellites, GPS/GNSS, etc.
    }
  }

\textbf{Response}: Modifications made on page 2.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: “The principal data source for forecasting of
      geomagnetic indices is the OMNI data set hosted by NASA...” This
      sentence is wrong. For Kp ground-based magnetometers are used
      and for Dst real- time data from L1 (from either the DSCOVR or
      ACE satellite) is used. 
    }
  }

\textbf{Response}: Modifications made on page 2 section 1.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: The data source for real-time forecasting is
      not OMNI. It is fine to discuss the OMNI data since that is what
      is used later in the chapter for the example study, but it might
      be better to discuss the OMNI data when working through the
      example so as not to confuse the reader about real- time data
      sources used for real-time space weather forecasting.
    }
  }

\textbf{Response}: It is mentioned on page 2, section 1.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: If you are only using 1-hour values, how can
      you get a useful real-time prediction?
    }
  }

\textbf{Response}: The question is not relevant to the scope of the
chapter, if one is interested in making predictions for geomagnetic
indices minutes in the future then one may apply the methodology
outlined to minute cadence data.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: Please define what m and K are within the text.
    }
  }

\textbf{Response}: Defined on page 4, section 3.1

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: Please clarify why you are using hourly OMNI
      data to predict Dst. Wouldn’t it be better to use the 1-minute
      data if you want to actually generate a useful forecast?
    }
  }

\textbf{Response}: This question is beyond the scope of this
chapter. The aim of the chapter is to educate practitioners on
applications of the GP methodology for time series prediction, the
choice of data cadence is not of primary relevance here. Depending on
the particular application and research questions being considered,
practitioners would choose which data set is more suitable for their purpose.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: When you mention the lag time p, more
      discussion would be helpful. How far back in time do you
      typically need to go? Does it depend more on the amount of time
      or the number of points? If I use 1-min data instead of 1-hour
      data do I need to use more points?
    }
  }

\textbf{Response}: Modifications made in section 5.4. Choice of lag
time p is empirical, how it changes depends on the characteristics of
the particular time series and data set being considered.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: Be consistent in either writing “sub-storm” or “substorm”
    }
  }

\textbf{Response}: Modifications made in the manuscript.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: Define RBE within the text in addition to in the table.
    }
  }

\textbf{Response}: RBF has been defined on page 7, section 5.2.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: Hyper-parameters is not hyphenated in the
      table while it is elsewhere within the text.
    }
  }

\textbf{Response}: Modifications made throughout the manuscript.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: Define Q in the text when it is first discussed.
    }
  }

\textbf{Response}: Changes made in section 5.3.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: In section 6.3 a more detailed description of
      how you got to the last equation would be useful (as would
      equation numbers).
    }
  }

\textbf{Response}: The equation is nothing but the logarithm of the
density function of the multivariate gaussian which is a very well
known equation in mathematics and science. (Section 5.3 page 8)

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: “Some of the techniques used for model
      selection in the context of GPR include.” This is a fragment of
      a sentence.
    }
  }

\textbf{Response}: Changes made on page 9.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: In the first paragraph of section 6.4, please
      explain why it would not be beneficial to group the quantities
      with the kernel hyper-parameters in the model selection procedure.
    }
  }

\textbf{Response}: Modifications made in section 5.4 on page 10.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: Why does Table 2 come after Tables 3 and 4 in
      the text?
    }
  }

\textbf{Response}: Appropriate corrections made in the manuscript.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: When choosing performance metrics, there are
      several other metrics that are traditionally used by the weather
      forecasting community that would be good to include such as the
      Heidke Skill Score (HSS). A great reference is Statistical
      Methods in Atmospheric Science by Daniel Wilks.
    }
  }

\textbf{Response}: The Heikde Skill Score is a metric used for
categorical predictive problems, a reference to the aforementioned
book has been added in section 6 point 3 on page 11.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: When discussing the maximum time lag, use
      units (min or hour) and discuss how these results would change
      if using minutes vs. hours.
    }
  }

\textbf{Response}: It is not possible to rigourously state how the
maximum time lag would change if using minutes instead of hours, as it
is empirically determined. That is why the chapter discusses a
workflow which can be applied in either case.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: Please elaborate on how the decision is made
      as to which model order yields the “best performance”.
    }
  }

\textbf{Response}: Changes have been made in section 6 point 4.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: This section could benefit from a more though
      discussion. As this is a text book, I think it would be useful
      for students if a more comprehensive description was included.
    }
  }

\textbf{Response}: Changes have been made to section 9.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: What did the persistence model predict? Could
      you plot that on the graph as well? A discussion of the
      persistence forecast results compared to these in the text would
      be helpful.
    }
  }

\textbf{Response}: Persistence model simply predicts the Dst value of
the previous hour as the prediction for the current hour. It is simply
a baseline predictor which can be used for simple evaluation of
predictive models.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: How much lead time do you get with this method? 
    }
  }

\textbf{Response}: Since the example is of one hour ahead prediction,
it gives a lead time of one hour.

}

\item{

\fbox{%
    \parbox{\textwidth}{%
      \textbf{Comment}: The chapter just ends. A conclusion or summary
      would be helpful.
    }
  }

\textbf{Response}: A conclusion has been added in section 9.

}

\end{enumerate}

\end{document}
